---
title: "Generalized Least Squares (GLSs)"
author: '"YOUR NAME HERE"'
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
require(nlme)
require(sp)
require(ape)
require(MASS)
```

# **The Set-up**

## *Learning objectives.*
1.	Implement as Generalized Least Squares model that considers Spatial correlation between observations in `R`. 
2.	Determine the fit of a Generalized Least Squares model.
3.	Reduce a Generalized Least Squares model to the minimum adequate number of predictors using a stepwise procedure.

## *Before you start*
The practical will be run using *RStudio cloud* to avoid individual machine troubleshooting. This means you **need** to have an account for *RStudio cloud*.

However, remember to have [`R`](https://cran.r-project.org) and [R-studio](http://www.rstudio.com/) in your/room computer so you can work on the project after class.

**You should have also **
1. Read chapter 4 of Zuur mixed effect models book.
1. Read chapter 26  of Crawley the R book.
2. Done the Tutorial for this week.

## *The way it's going to be run*

In this 3h practical you will to implement what you have learned in the class about Regression Analyses in `R`.

You will be building and **R-Markdown** document during the class. **R-Markdown** documents allows you to combine your code, its results, and your prose commentary into a single document. If you manage to produce a document your code works. The resulting file (a html) is what you will hand-in at the end of the practical.

This 3h practical is divided into seven parts (24 tasks in total). Also try to write your own `R`-code and run it **DO NOT COPY THE CODE OF SOMEONE ELSE**. Writing your code and figuring out why it does not work is the best way to understand the dynamics of the `R` language and the tasks you are doing. Also, you have an excellent teaching assistant resource - ask him and/or me all the questions you have.

The final document will have three classes of boxes

<div class="alert alert-info">
<p class="text-center"; style = "font-size: 20px">
The text in the blue filled box marked as <strong>Your task</strong> states what you need to do
</p>
</div>

```{r }
#
# The code in the window marked like this shows where you need to write your code
#

```

<div class="alert alert-success">
<p class="text-center"; style = "font-size: 20px">
The text in the green filled box marked as <strong>Question</strong> phrases a question, and where you will place the answer to the <strong>question</strong> is marked as **PLACE YOUR ANSWER HERE**. </p>
</div>

## About the data you will use.

The data to be used in this week practical is a combination of the biodiversity data  from the [Biowide](https://ecos.au.dk/forskningraadgivning/temasider/biowide) project, which aimed at collecting biodiversity data from 130 terrestrial sampling sites across Denmark. The dataset contains data for Plants (vascular plants and bryophytes), fungi (lichens and macrofungi), gastropods, and arthropods. You can get the [biodiversity data](https://www.gbif.org/dataset/cd4eb3ec-0a18-42b4-bda4-155716ddd7b1#description) from the Danish Biodiversity Information Facility. For further information on the project check the publication.

Although environmental variables were collected as part of Biowide, this data is not available for our use. The data on environmental variables included in the table you will use comes from multiple Geospatial datasets I complied for this course.

Here you will focus on the flowing variables:

**1) decimalLongitude**: East-West position in decimal-degrees of a sample site coming from [Biowide](https://ecos.au.dk/forskningraadgivning/temasider/biowide).

**2) decimalLatitude**: North-South position in decimal-degrees of a sample site coming from [Biowide](https://ecos.au.dk/forskningraadgivning/temasider/biowide). 

**3) Log.S.AllGrp**: The total species richness of a sample site (sum of the unique species of all sampled groups in a site) coming from [Biowide](https://ecos.au.dk/forskningraadgivning/temasider/biowide).

**4) AgrBasMp**: Density of agricultural land cover types as defined by [BASEMAP](https://envs.au.dk/en/research-areas/society-environment-and-resources/land-use-and-gis/basemap).

**5) WetLndBasMp**: Density of Wetlands cover types as defined by [BASEMAP](https://envs.au.dk/en/research-areas/society-environment-and-resources/land-use-and-gis/basemap).

**6) Slope30mAgg**: the maximum rate of change in value from that cell to its neighbours estimated based on the [SRTM-30m DEM](https://www2.jpl.nasa.gov/srtm/).

**7) NDVI**: Normalized difference vegetation index, a simple graphical indicator that can be used to analyse vegetation greenness and has strong relation with productivity [Copernicus Global Land Service](https://land.copernicus.eu/global/products/ndvi).

**8) Mean Annual Temperature (MAT)**: Annual average of temperatures  coming the [CHELSA-climatologies](https://chelsa-climate.org) ..

**9) Dis2Cost**: Distance in km of a sampled site to the coast estimated by me using Denmark high resolution contour map coming form the Database of Global Administrative Areas [GADAM](https://gadm.org).

**10) HII**: [Human Influence Index](https://sedac.ciesin.columbia.edu/data/set/wildareas-v2-human-influence-index-geographic), a measurement of the anthropogenic impacts on the environment.

**11) PrecSea**: Seasonality of precipitation measured as the variability of monthly values (using the CV) coming the [CHELSA-climatologies](https://chelsa-climate.org).


## 1) The first step: Loading the data

<div class = "alert alert-info">
**Your task:**

Load the data in *Biowide_AllSppRichV2.csv*, located in the files of the project

Use `read.csv` to load the file - The file is comma-separated

Save the object as an object named `DK.Biodiv`.
</div>

```{r LoadData}
# Read the file and save it as an object named `DK.Biodiv`

DK.Biodiv <-  

# Use the function head to print the five first rows

```

## 2) Building a regression.

<div class = "alert alert-info">
**Your task:**

Use the function `lm()` to build  model where all **continuous** predictors, with the exception of `decimalLongitude` and `decimalLatitude` are used to predict `Log.S.AllGrp`.

</div>
  
```{r glmMod1}
# Build a lm to predict Log.S.AllGrp based on a linear combination of all NON-positional predictors
DK.Biodiv.lm <- 
  
# Print the summary of DK.Biodiv.lm

```

## 3) Check your regression.

<div class = "alert alert-info">
**Your task:**

* Evaluate graphically the assumptions of your full linear model (`DK.Biodiv.lm`).
</div>
  
```{r glmModDiagFig, fig.dim=c(6,6)}
# Evaluate graphically the assumptions of your full linear model

```

<div class="alert alert-success">
**Question:** Are the following assumptions meet??

1) Homogeneity of variances?
**PLACE YOUR ANSWER HERE**

2) Normality of residuals?
**PLACE YOUR ANSWER HERE**

3) Observations with large leverage?
**PLACE YOUR ANSWER HERE**
</div>

<div class = "alert alert-info">
**Your task:**

* Evaluate if the predictors in `DK.Biodiv.lm` are collinear. For this, use the function `vif()` for the package `car`.
</div>
  
```{r glmModcollinear}
# Evaluate if the predictors in `DK.Biodiv.lm` are collinear.

```

<div class="alert alert-success">
**Question:** Are predictors collinear?.

**PLACE YOUR ANSWER HERE**
</div>

<div class = "alert alert-info">
**Your task:**

* Evaluate graphically the assumptions of independence of observations. (`DK.Biodiv.lm`). For this:
    1) Plot the Residuals vs Each predictor.
    2) Add a smoother trend line using the function `lowess()`.
</div>

```{r glmModIndepen, fig.dim=c(6,10)}
# Plot the Residuals vs the predictor

# add a smoother trend line using lowess

```

<div class="alert alert-success">
**Question:** Is there independence of observations?.

**PLACE YOUR ANSWER HERE**
</div>


## 4) Detecting spatial autocorrelation

<div class = "alert alert-info">
**Your task:**

* Create a Spatial object with the data in `DK.Biodiv`. For this, you will:
    1) Copy the data in `DK.Biodiv` into an object names `DK.Biodiv.Spatial`.
    2) Extract the standardized residuals of the `DK.Biodiv.lm` object using the function `rstandard()`; and add these to `DK.Biodiv.Spatial.` as a variable named `Resid`.
    3) Transform the `DK.Biodiv.Spatial` object from a `data.frame` to a `SpatialPointsDataframe` using the function `coordinates()`.
    4) Define the right reference system to `DK.Biodiv.Spatial` using the function `proj4string()`..
    5) Re-project `DK.Biodiv.Spatial` to `UTM/ETRS89` using the function `spTransform()`.
 
 </div>
```{r SpatialObj}
# Copy the data in `DK.Biodiv` into an object names `DK.Biodiv.Spatial`.
DK.Biodiv.Spatial <- DK.Biodiv

# Transform the `DK.Biodiv.Spatial` to a `SpatialPointsDataframe`

# Define the right reference system to `DK.Biodiv.Spatial`.

# Re-project `DK.Biodiv.Spatial` to `UTM/ETRS89`.

# Add the projected X and Y to the Data frame of the Spatial object
DK.Biodiv.Spatial$X <- 
DK.Biodiv.Spatial$Y <- 

# Plot the spatial object you created

# Load a map of Denmark
DK.Map <- readRDS("data/gadm36_DNK_0_sp.rds")
# Re-project `DK.Map` to `UTM/ETRS89`.
DK.Map <- 
# plot DK.Map over DK.Biodiv.Spatial 

```

<div class = "alert alert-info">
**Your task:**

* Generate a map to represent the spatial variation in the residual variation from the `DK.Biodiv.lm` model. For this, you will:
    1) Add to `DK.Biodiv.Spatial` the normalized residuals from the `DK.Biodiv.lm` model as a variable names `Resid`. You will get these using the function `rstandard()`.
    2) Use the function `bubble()` from the package `sp`to generate this spatial representation of the normalized residual variation.
    
 </div>
 
```{r bubblePlot}
# Add to `DK.Biodiv.Spatial` the normalized residuals to the Spatial object
DK.Biodiv.Spatial$Resid <- 

# Spatial representation of the normalized residual variation.

```
 

<div class="alert alert-success">
**Question:** there is a clear spatial pattern in the residuals?.

**PLACE YOUR ANSWER HERE**

</div>


## 5) Building a semi-variogram to quantify the Autocorrelation.

<div class = "alert alert-info">
**Your task:** 

* Re-run your best model using a Generalised least Squares method. For this, you will use the function `gls()` and the data in `DK.Biodiv.Spatial`.

* Save the the model as an object named `DK.Biodiv.gls`.

</div>

```{r GLSmod}
# Make the reduced model a Generalised least Squares model
DK.Biodiv.gls <- 

# call the summary of DK.Biodiv.gls

```

<div class="alert alert-success">
**Questions:** 

1) Is there any change in the regression coefficients when ruing the model with the function `gls()`?.

**PLACE YOUR ANSWER HERE**

2) Why you do not get a R^2^ or Adjusted-R^2^ when ruining the model with the function `gls()`?

**PLACE YOUR ANSWER HERE**

</div>

<div class = "alert alert-info">
**Your task:**

* Generate a variogram for the residuals from `DK.Biodiv.gls`. For this, you will use the function `Variogram()` from the `nlme` package. See the help file from the function `Variogram()` to see how to use this. 
</div>

```{r Semivar}
# Generate a semi-variogram for the residuals from `DK.Biodiv.gls`. 

DK.Biodiv.gls.Vargrm <- 
  
# Plot the semi-variogram

```

<div class="alert alert-success">
**Question:** Is there autocorrelation in the residuals?.

**PLACE YOUR ANSWER HERE**

</div>

## 6) Moran-I: An Statistic for Autocorrelation.


<div class = "alert alert-info">
**Your task:** 

* Estimate the Moran's I statistic (which is a measure of spatial autocorrelation) using the function `Moran.I()` of the package `ape`. For this, you will:
    1) Create a distance matrix between all the observations using the function `dist()`, and the projected coordinates. Save the distance matrix as an object named `DistDKBio`.
    2) Extract the `normalized` residuals of the gls model (`DK.Biodiv.gls`). Save these as a object names `NorResGLS`. 
    3) Estimate Moran I of the residuals using the function `Moran.I()` of the package `ape`.

</div>

```{r MoranI}
# Create a distance matrix using the function dist() on the coordinates of the sampled locations
DistDKBio <- 

#Extract the `normalized` residuals
NorResGLS <- 

# Estimate Moran I of the residuals
Moran.I(x = , # the variable to evaluate 
        w = as.matrix(DistDKBio) # the Distance matrix 
        )
```

<div class="alert alert-success">
**Question:** Is there autocorrelation in the residuals based on Moran I? - Check the help file of the function `Moran.I()` to see wit is tested.

**PLACE YOUR ANSWER HERE**
</div>

## 7) Accommodating spatial autocorrelation in the model.

<div class = "alert alert-info">
**Your task:** 

* Refit the model `DK.Biodiv.gls` using the flowing correlation structures:
  1) Exponential. 
  2) Gaussian
  3) Spherical
  4) Rational quadratic
  
Remember to use the projected spatial object, and use the projected Coordinates when defining the correlation structures.
</div>

```{r Cor Str}
# Fit an Exponential correlation model
DK.Biodiv.glsExp <- 

# Fit a Gaussian correlation model
DK.Biodiv.glsGaus <- 
  
# Fit a Rational quadratic correlation model
DK.Biodiv.glsRatio <- 

# Fit an Spherical correlation model
DK.Biodiv.glsSpher <- 
  
```

<div class = "alert alert-info">
**Your task:** 
* Contrast the AIC of the different models
</div>


```{r AICcontrast}
#Contrast the AIC of the different models
AIC(, # Exponential model
    , # Gaussian model
    , # Ratio model
    r # Sperical model
    )
```

<div class="alert alert-success">
**Question:** Based on the AIC which is the best model?.

**PLACE YOUR ANSWER HERE**

</div>

<div class = "alert alert-info">
**Your task:** 
Contrast the  the best model with a correlation structure to that of the model with out one (`DK.Biodiv.gls`). For this, you will use the function `anova()`. 
</div>

```{r finalcontrast}
#Contrast the  the best model with a correlation structure to that of the model with out one

```

<div class="alert alert-success">
**Question:** Best on the AIS which is the best model .

**PLACE YOUR ANSWER HERE**
</div>

## 8) Select the best Fixed effects.

<div class = "alert alert-info">
**Your task:** 

* Now that you have defined the best correlation structure; select the best set of predictors. For this, you will:
  1) Redefine the model with a correlation structure, by updating the inferencer method. Use the function `update()` for this.
  2) Use the function `setpAIC()` from the package `MASS` to do the model section using a `both` directions stepwise search.
  3) Redefine the model with a correlation structure, by updating the inferencer method. Use the function `update()` for this.

</div>

```{r ModelSel}
# updating the inferencer method of the model with a correlation structure
DK.Biodiv.glsSpher.ML <-

# do the model section using a `both` directions stepwise search.
DK.Biodiv.glsSpher.ML.Updated <- 

# updating the inferencer method of the model with a correlation structure
DK.Biodiv.glsSpher.ML.Updated <- 

  # Call the summary of the reduced model after updating the inference method.

```

## 9) Last check

<div class = "alert alert-info">
**Your task:**

* Evaluate graphically the assumptions of your reduced linear model. Here only focus on:
  1) Homogeneity of variances.
  2) Normality of the residuals

</div>
  
```{r glsModDiagFig}
# Evaluate graphically the assumption of Homogeneity of variances in your reduced gls model with a correlation structure


# Evaluate graphically the assumption of normality of residuals in your reduced gls model with a correlation structure

```

<div class="alert alert-success">
**Question:** Are the following assumptions meet??

1) Homogeneity of variances?
**PLACE YOUR ANSWER HERE**

2) Normality of residuals?
**PLACE YOUR ANSWER HERE**

</div>

<div class = "alert alert-info">
**Your task:**

* Generate a variogram for the residuals from the final reduced model. For this, you will use the function `Variogram()` from the `nlme` package.
</div>

```{r SemivarFinal}
# Generate a semi-variogram for the residuals from the reduced model. 

DK.Biodiv.gls.Vargrm2 <- 

# Plot the semi-variogram

```


<div class="alert alert-success">
**Question:** Is there autocorrelation in the residuals?.

**PLACE YOUR ANSWER HERE**

</div>