---
title: "Generalized Linear Models (GLMs)"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#libraries
library(learnr)
DK.Biodiv <-  read.csv("data/Biowide_AllSppRich.csv")
DK.Biodiv.glm <- glm(S.AllGrp ~ NatDensBasMp + Dis2Cost + HII + Slope30mAgg + PrecSea, #  the Equations defining the response and additive combination of predictors
                     data = DK.Biodiv, # specify the Object where the data is in
                     family = "poisson")
Summ.DK.Biodiv.glm <- summary(DK.Biodiv.glm)

DK.Biodiv.glmQuasi <- glm(S.AllGrp ~ NatDensBasMp + Dis2Cost + HII + Slope30mAgg + PrecSea, #  the Equations defining the response and additive combination of predictors
                          data = DK.Biodiv, # specify the Object where the data is in
                          family = "quasipoisson")

DK.Biodiv.glmnb <- MASS::glm.nb(S.AllGrp ~ NatDensBasMp + Dis2Cost + HII + Slope30mAgg + PrecSea, #  the Equations defining the response and additive combination of predictors
                                data = DK.Biodiv) # specify the Object where the data is in
Const.glm.nb <- MASS::glm.nb(S.AllGrp ~ 1, #  the Equations of a Constant model
                             data = DK.Biodiv # specify the Object where the data is in
)
DK.Biodiv.glmnb.Back <- step(DK.Biodiv.glmnb,
                             direction = "backward")

```

# **The Set-up**

## *Learning objectives.*
1.	Implement as simple/multiple Generalized Linear Models in `R`. 
2.	Determine the fit of a Generalized Linear Models and establish how good a model is.
3.	Reduce a Generalized Linear Models to the minimum adequate number of predictors using a stepwise procedure.

## About the data you will use.

The data to be used in this week practical is a combination of the biodiversity data  from the [Biowide](https://ecos.au.dk/forskningraadgivning/temasider/biowide) project, which aimed at collecting biodiversity data from 130 terrestrial sampling sites across Denmark. The dataset contains data for Plants (vascular plants and bryophytes), fungi (lichens and macrofungi), gastropods, and arthropods. You can get the [biodiversity data](https://www.gbif.org/dataset/cd4eb3ec-0a18-42b4-bda4-155716ddd7b1#description) from the Danish Biodiversity Information Facility. For further information on the project check the publication.

Although environmental variables were collected as part of Biowide, this data is not available for our use. The data on environmental variables included in the table you will use comes from multiple Geospatial datasets I complied for this course.

Here you will focus on the flowing variables:

**1) decimalLongitude**: East-West position in decimal-degrees of a sample site coming from [Biowide](https://ecos.au.dk/forskningraadgivning/temasider/biowide).

**2) decimalLatitude**: North-South position in decimal-degrees of a sample site coming from [Biowide](https://ecos.au.dk/forskningraadgivning/temasider/biowide). 

**3) S.AllGrp**: The total species richness of a sample site (sum of the unique species of all sampled groups in a site) coming from [Biowide](https://ecos.au.dk/forskningraadgivning/temasider/biowide).

**4) landsdele**: Regional political unit of Denmark coming form the Database of Global Administrative Areas [GADAM](https://gadm.org).

**5) Geology**: Geological formation coming from the [Denamrk's geological service](https://eng.geus.dk/products-services-facilities/data-and-maps/maps-of-denmark).

**6) NatDensBasMp**: Density of natural land cover types as defined by [BASEMAP](https://envs.au.dk/en/research-areas/society-environment-and-resources/land-use-and-gis/basemap).

**7) Dis2Cost**: Distance in km of a sampled site to the coast estimated by me using Denmark high resolution contour map coming form the Database of Global Administrative Areas [GADAM](https://gadm.org).

**8) HII**: [Human Influence Index](https://sedac.ciesin.columbia.edu/data/set/wildareas-v2-human-influence-index-geographic), a measurement of the anthropogenic impacts on the environment.

**9) Slope30mAgg**: the maximum rate of change in value from that cell to its neighbours estimated based on the [SRTM-30m DEM](https://www2.jpl.nasa.gov/srtm/).

**10) PrecSea**: Seasonality of precipitation measured as the variability of monthly values (using the CV) coming the [CHELSA-climatologies](https://chelsa-climate.org) .


## The first step: Loading the data

<div class = "alert alert-info">
**Your task:**

Load the data in *Biowide_AllSppRich.csv*, located in the files of the project

Use `read.csv` to load the file - The file is comma-separated

Save the object as an object named `DK.Biodiv`.
</div>

```{r LoadData, exercise=TRUE, exercise.lines=15} 
# Read the file and save it as an object named `DK.Biodiv`
DK.Biodiv <-  -------("-------")
# Use the function head to print the five first rows
-------(-------)
```
```{r LoadData-solution}
# Read the file and save it as an object named `DK.Biodiv`
DK.Biodiv <-  read.csv("data/Biowide_AllSppRich.csv")
# Use the function head to print the five forst rows
head(DK.Biodiv)
```

## Data Exploration

<div class="alert alert-info">
**Your task:**

Using the `DK.Biodiv` data, you will now build a box plot for each continuous predictor variable. You will also add a "title" to each plot using the `main` argument.

The continuous predictors are:
1) *Nature Density* [`NatDensBasMp`]
2) *Distance to coast* [`Dis2Cost`]
3) *Human Impact Index* [`HII`]
4) *Slope* [`Slope30mAgg`]
5) *Precipitation Seasonality* [`PrecSea`]

</div>
```{r Boxplot1, exercise=TRUE, exercise.lines=15} 
# Set-up the plotting space
par(mfrow=c(2,3))

# Box plot for Nature Density
boxplot(-------$-------,
        main = "Nature Density")
# Box plot for Distance to coast 
boxplot(-------$-------,
        main = "Distance to coast")
# Box plot for Human Impact Index
boxplot(-------$-------,
        main = "Human Impact Index")
# Box plot for Slope
boxplot(-------$-------,
        main = "Slope")
# Box plot for Precipitation Seasonality
boxplot(-------$-------,
        main = "Precipitation Seasonality")
```
```{r Boxplot1-solution}
# Set-up the plotting space
par(mfrow=c(2,3))

# Box plot for Nature Density
boxplot(DK.Biodiv$NatDensBasMp,
        main = "Nature Density")
# Box plot for Distance to coast 
boxplot(DK.Biodiv$NatDensBasMp,
        main = "Distance to coast")
# Box plot for Human Impact Index
boxplot(DK.Biodiv$HII,
        main = "Human Impact Index")
# Box plot for Slope
boxplot(DK.Biodiv$Slope30mAgg,
        main = "Slope")
# Box plot for Precipitation Seasonality
boxplot(DK.Biodiv$PrecSea,
        main = "Precipitation Seasonality")
```


<div class="alert alert-success">
**Question**: For which variable can you detect outlier observation.

**WRITE YOUR ANSWER HERE**
</div>


<div class = "alert alert-info">
**Your task:**
Estimate the pairwise correlation between variables (using the `cor()` function) and plotting their values using the function `image.plo()` of the package `fields`.

</div>

```{r Colin1, exercise=TRUE, exercise.lines=15} 
# Assess the correlation between predictors and save this as an object named Cor.pred.DK.Biodiv
Cor.pred.DK.Biodiv <- -------(-------[, c("-------",
                                          "-------",
                                          "-------",
                                          "-------",
                                          "-------")], 
        method = "-------")
#Call the Cor.pred.DK.Biodiv object
-------
# Use the  function image.plo() of the package fields to generate a plot of the correlation matrix
fields::image.plot(-------,
                   zlim=c(-1,1), # Sets the range of values to get colors from between -1 and1
                   col = hcl.colors(100,"RdYlBu") # sets the colors to a Red-Yellow-Blue Palette.
                   )
```
```{r Colin1-solution}
# Assess the correlation between predictors and save this as an object named Cor.pred.DK.Biodiv
Cor.pred.DK.Biodiv <- cor(DK.Biodiv[, c("NatDensBasMp", "Dis2Cost", "HII", "Slope30mAgg", "PrecSea")], 
        method = "spearman")
#Call the Cor.pred.DK.Biodiv object
Cor.pred.DK.Biodiv
# Use the  function image.plo() of the package fields to generate a plot of the correlation matrix
fields::image.plot(Cor.pred.DK.Biodiv,
      zlim=c(-1,1), # Sets the range of values to get colors from between -1 and1
      col = hcl.colors(100,"RdYlBu") # sets the colors to a Red-Yellow-Blue Palette.
      )
```

<div class="alert alert-success">
**Question:** Which pairs of predictors have a correlation above the 0.7 cut-off?

**WRITE YOUR ANSWER HERE**

</div>


<div class = "alert alert-info">
**Your task:**

With the data in memory (saved as an object named `DK.Biodiv`), you will evaluate the multiple-collinearity of the predictor variables [`NatDensBasMp`, `Dis2Cost`, `HII`, `Slope30mAgg`, `PrecSea`] using the variance inflation index. For this you will:

* Build a linear model where`S.AllGrp` is the response variable, and `NatDensBasMp`, `Dis2Cost`, `HII`, `Slope30mAgg`, and `PrecSea` are set as linear predictors with NO interactions.

* Use a the function `vif()` of the `car` package to estimate the variance inflation index.

</div>

```{r Colin2, exercise=TRUE, exercise.lines=15} 
# Build a linear model where`S.AllGrp` is the response variable, and `NatDensBasMp`, `Dis2Cost`, `HII`, `Slope30mAgg`, and `PrecSea` are set as linear predictors with NO interactions.
Colin.Model <- -------(------- ~ ------- + ------- + ------- + ------- + -------,
                  data = -------)
# Use a the function `vif()` of the `car` package to estimate the variance inflation index. 
Colin.DK.Pred <- car::vif(mod = -------) #
# Call the Colin.DK.Pred object
-------

# Estimating each predictor tolerance (1/vif)
-------

```
```{r Colin2-solution}
# Build a linear model where`S.AllGrp` is the response variable, and `NatDensBasMp`, `Dis2Cost`, `HII`, `Slope30mAgg`, and `PrecSea` are set as linear predictors with NO interactions.
Colin.Model <- lm(S.AllGrp ~ NatDensBasMp + Dis2Cost + HII + Slope30mAgg + PrecSea,
                  data = DK.Biodiv)
# Use a the function `vif()` of the `car` package to estimate the variance inflation index. 
Colin.DK.Pred <- car::vif(mod = Colin.Model) #
# Call the Colin.DK.Pred object
Colin.DK.Pred

# Estimating each predictor tolerance (1/vif)
1/Colin.DK.Pred

```
<div class="alert alert-success">
**Question:** Do you see collinearity between the predictors?.

**WRITE YOUR ANSWER HERE**
</div>

## How does my response variable change as a function of each predictor?

<div class = "alert alert-info">
**Your task:**

Graphically explore how `S.AllGrp` changes as a function of either `NatDensBasMp`, `Dis2Cost`, `HII`, `Slope30mAgg`, `PrecSea`. For this, you will:

* Plot a scatter plots for each `S.AllGrp ~` *predictor* combination.

* Build a linear model using the function `?line()`.

* Plot a line representing the association between `S.AllGrp` **AND** *predictor*.

* Using the function `legend()`, add to the plot the person correlation coefficient of the relation and the P-value, calculated with the function `cor.test()`, .

**YOU WILL USE a `for()`loop to make this figure**.

</div>
  
```{r UnivarRel1, exercise=TRUE, exercise.lines=20,fig.dim=c(6,9)} 
# Start by setting a plotting space with three (3) rows and two (2) columns, making sure that plots are added row-wise
par(mfrow = c(3, 2)) # add a vector with the number of rows and number of columns in the plotting space
# start a for() Loop
for (Var in c("NatDensBasMp", "Dis2Cost", "HII", "Slope30mAgg", "PrecSea")){ 
# Scatter plot of S.AllGrp vs the evaluated predictor using the test.
  
 plot(x = -------, # The values for the predictor
      y = -------, # The values for the response, 
      pch = -------, # make the points filled circles   
      main = -------,  # Add a title to the figure
      ylab = "Species Richness", # Y-axis name
      xlab = ------- # X-axis name
      )

 # To  add a trend line you first need a regression object created with the line() function.
  Reg.Tmp <- -------(x = -------, # The values for the predictor
                     y = -------) # The values for the response, )
# Add a trend line using the abline() function on the lm function created above
  -------(-------, ## the line object to plot
          col = "red", # the colour of the line
          lwd = 2,  # the width of the line
          lty = 2) # the type of the line

# Estimate the correlation between S.AllGrp and the predictor.
 Cor.Tmp <- -------(x = -------),
                    y = -------)

# Add it to the figure using the function legend  
  legend("topright",
         legend = paste0("r = ",
                         -------,
                         "; p = ",
                         -------)
         )
  }
```
```{r UnivarRel1-solution}
# Start by setting a plotting space with three (3) rows and two (2) columns, making sure that plots are added row-wise
par(mfrow = c(3, 2)) # add a vector with the number of rows and number of columns in the plotting space
# start a for() Loop
for (Var in c("NatDensBasMp", "Dis2Cost", "HII", "Slope30mAgg", "PrecSea")){ 
# Scatter plot of S.AllGrp vs the evaluated predictor using the test.
  
 plot(x = DK.Biodiv[,Var], # The values for the predictor
      y = DK.Biodiv[,"S.AllGrp"], # The values for the response, 
      pch = 19, # make the points filled circles   
      main = Var,  # Add a title to the figure
      ylab = "Species Richness", # Y-axis name
      xlab = Var # X-axis name
      )

 # To  add a trend line you first need a regression object created with the line() function.
  Reg.Tmp <- line(x = DK.Biodiv[,Var], # The values for the predictor
                  y = DK.Biodiv[,"S.AllGrp"]) # The values for the response, )
# Add a trend line using the abline() function on the lm function created above
  abline(Reg.Tmp, ## the line object to plot
         col = "red", # the colour of the line
         lwd = 2,  # the width of the line
         lty = 2) # the type of the line

# Estimate the correlation between S.AllGrp and the predictor.
 Cor.Tmp <- cor.test(x = log10(DK.Biodiv[,Var]),
                     y = DK.Biodiv[,"S.AllGrp"])
 
  legend("topright",
         legend = paste0("r = ",
                         round(Cor.Tmp$estimate,3),
                         "; p = ",
                         round(Cor.Tmp$p.value,3)))
summary(lm(formula(paste0("S.AllGrp~",Var)),
   data=DK.Biodiv))
  
  }
```

<div class="alert alert-success">
**Question:** Based on these graphical representations, what do you see?

**WRITE YOUR ANSWER HERE**
</div>


## Why do you need a GLM?	

Answer the question below based on what you discussed in the theory session. 

<div class="alert alert-success">
**Question:** Which is the family of the model to will use to predict `S.AllGrp`?.
  
**WRITE YOUR ANSWER HERE**
</div>

## GLM and hypothesis testing

When the response variable is an integer (i.e. Counts of individuals), you model $\lambda_(x)$, that is, the mean count ($\lambda$) given value of $X$ or a combination of $X_i$. The usual model fitted to such data is a Poisson regression model.

The model looks like this:
$g(x) = \lambda_(x) = \beta_0 + (\beta_1*NatDensBasMp) + (\beta_2*Dis2Cost) + (\beta_3*HII) + (\beta_4*Slope30mAgg) + (\beta_5*PrecSea)$

Also, remember that the error terms from the poisson model are generally not normality distributed; because the response variable is discrete, the error terms have a Poisson distribution. Also there is no homogeneity of variances, as the variance is the same as the mean in a Poisson regression.

<<div class="alert alert-success">
**Question:** Based on the regression formulation described in the point above, which is/are the H~0~’s you are interested in testing?

**WRITE YOUR ANSWER HERE**
</div>
  
## Building a Poisson regression.

<div class = "alert alert-info">
**Your task:**

Use the `glm()` function to evaluate the proposed poisson regression. Remember to include the adequate error `family` for the type of data in `S.AllGrp`.
  
</div>
  
```{r glmMod, exercise=TRUE, exercise.lines=15} 
# Build a glm to predict S.AllGrp based on a linear combination of NatDensBasMp, Dis2Cost, HII, Slope30mAgg, PrecSea
# Save the Model as an object named DK.Biodiv.glm
DK.Biodiv.glm <- -------(------- ~ ------- + ------- + ------- + ------- + -------, #  the Equations defining the response and additive combination of predictors
                  data = -------, # specify the Object where the data is in
                  family = "-------") # Argument defining the adequate error `family` given the type of the response variable (quasi)binomial for presence/absence and proportion data, and (quasi)Poisson for counts.
-------
```
```{r glmMod-solution}
# Build a glm to predict S.AllGrp based on a linear combination of NatDensBasMp, Dis2Cost, HII, Slope30mAgg, PrecSea
# Save the Model as an object named DK.Biodiv.glm
DK.Biodiv.glm <- glm(S.AllGrp ~ NatDensBasMp + Dis2Cost + HII + Slope30mAgg + PrecSea, #  the Equations defining the response and additive combination of predictors
                  data = DK.Biodiv, # specify the Object where the data is in
                  family = "poisson") # Argument defining the adequate error `family` given the type of the response variable (quasi)binomial for presence/absence and proportion data, and (quasi)Poisson for counts.

DK.Biodiv.glm
```

## Initial summary of the regression

<div class = "alert alert-info">
**Your task:**

Build a `data.frame` that contains the maximum likelihood estimates of the model parameters. The `data.frame` set-up should be as the table below. Also, this table needs to build by extracting the adequate elements from the `DK.Biodiv.glm` object, **NOT BY TYPING VALUES IN THE TABLE!**

+-----------------------+---------------+---------------+----------------+----------------+
| PARAMETER             | Estimate      | SE            | Z-value        | $p$            |
+=======================+===============+===============+================+================+
| $\beta_0$             |               |               |                |                |
+-----------------------+---------------+---------------+----------------+----------------+
| $\beta_{NatDensBasMp}$|               |               |                |                |
+-----------------------+---------------+---------------+----------------+----------------+
| $\beta_{Dis2Cost}$    |               |               |                |                |
+-----------------------+---------------+---------------+----------------+----------------+
| $\beta_{HII}$         |               |               |                |                |
+-----------------------+---------------+---------------+----------------+----------------+
| $\beta_{Slope30mAgg}$ |               |               |                |                |
+-----------------------+---------------+---------------+----------------+----------------+
| $\beta_{PrecSea}$     |               |               |                |                |
+-----------------------+---------------+---------------+----------------+----------------+
</div>

```{r glmmCoff, exercise=TRUE, exercise.lines=15} 
# use the summary() function on an object created using the glm() function to get the overview of the regression coefficients. 
# save this as an object named  Summ.DK.Biodiv.glm
Summ.DK.Biodiv.glm <- -------(-------)

# Call the Summ.DK.Biodiv.glm object
Summ.DK.Biodiv.glm
-------

# Now extract the coefficients from Summ.DK.Biodiv.glm with the coef() function. save this as an object named Coef.DK.Biodiv.glm
Coef.Summ.DK.Biodiv.glm <- -------(-------)
# print Coef.Summ.DK.Biodiv.glm, but round all numbers to 3 decimal points
-------(x = -------, # What to round?
      digits = -------) #number of decimal points
```
```{r glmmCoff-solution}
# use the summary() function on an object created using the glm() function to get the overview of the regression coefficients. 
# save this as an object named  Summ.DK.Biodiv.glm
Summ.DK.Biodiv.glm <- summary(DK.Biodiv.glm)
Summ.DK.Biodiv.glm

# Now extract the coefficients from Summ.DK.Biodiv.glm with the coef() function. save this as an object named Coef.DK.Biodiv.glm
Coef.Summ.DK.Biodiv.glm <- coef(Summ.DK.Biodiv.glm)
# print Coef.Summ.DK.Biodiv.glm, but round all numbers to 3 decimal points
round(x = Coef.Summ.DK.Biodiv.glm, # What to round?
      digits = 3) #number of decimal points
```

## Do you have an overdispersion problem?

<div class = "alert alert-info">
**Your task:**

Use the function summary on the `DK.Biodiv.glm` object created using the `glm()` function to compare the ratio between the residual deviance and the degrees of freedom.

</div>
  
```{r Overdisp, exercise=TRUE, exercise.lines=15} 
# use the summary() function on an object created using the glm() function to get the overview of the regression coefficients. 
# save this as an object named  Summ.DK.Biodiv.glm
Summ.DK.Biodiv.glm <- -------(-------)
# print Summ.DK.Biodiv.glm
Summ.DK.Biodiv.glm
```
```{r Overdisp-solution}
# use the summary() function on an object created using the glm() function to get the overview of the regression coefficients. 
# save this as an object named  Summ.DK.Biodiv.glm
Summ.DK.Biodiv.glm <- summary(DK.Biodiv.glm)
# print Summ.DK.Biodiv.glm
Summ.DK.Biodiv.glm
```

<div class="alert alert-success">
**Question:** Do you see an over-dispersion or under-dispersion problem?; if so, how you will solve the problem?

**WRITE YOUR ANSWER HERE**
</div>

## Addresing the over-dispersion problem.	

<div class = "alert alert-info">
**Your task:**

Use the `glm()` function to build a regression that could address the over-dispersion problem. Remember to include the adequate error `family` for the type of data in `S.AllGrp`.
  
</div>
  
```{r glmModQuasi, exercise=TRUE, exercise.lines=15} 
# Build a glm to predict S.AllGrp based on a linear combination of NatDensBasMp, Dis2Cost, HII, Slope30mAgg, PrecSea
# Save the Model as an object named DK.Biodiv.glm
DK.Biodiv.glmQuasi <- -------(------- ~ ------- + ------- + ------- + ------- + -------, #  the Equations defining the response and additive combination of predictors
                          data = -------, # specify the Object where the data is in
                          family = "-------") # Argument defining the adequate error `family` given the type of the response variable (quasi)binomial for presence/absence and proportion data, and (quasi)Poisson for counts.
# Call the summary() of the DK.Biodiv.glmQuasi object
-------(-------)
```
```{r glmModQuasi-solution}
# Build a glm to predict S.AllGrp based on a linear combination of NatDensBasMp, Dis2Cost, HII, Slope30mAgg, PrecSea
# Save the Model as an object named DK.Biodiv.glm
DK.Biodiv.glmQuasi <- glm(S.AllGrp ~ NatDensBasMp + Dis2Cost + HII + Slope30mAgg + PrecSea, #  the Equations defining the response and additive combination of predictors
                          data = DK.Biodiv, # specify the Object where the data is in
                          family = "quasipoisson") # Argument defining the adequate error `family` given the type of the response variable (quasi)binomial for presence/absence and proportion data, and (quasi)Poisson for counts.
# Call the summary() of the DK.Biodiv.glmQuasi object
summary(DK.Biodiv.glmQuasi)
```

<div class="alert alert-success">
**Question:** Is there an over-dispersion problem in this quasi-Poisson model?

**WRITE YOUR ANSWER HERE**
</div>


<div class = "alert alert-info">
**Your task:**

Use the `glm.nb()` function of the `MASS` package to build a negative-binomial regression that could address the over-dispersion problem. Remember to include the adequate error `family` for the type of data in `S.AllGrp`.
  
</div>
```{r glmNegBin, exercise=TRUE, exercise.lines=15} 
# Build a glm to predict S.AllGrp based on a linear combination of NatDensBasMp, Dis2Cost, HII, Slope30mAgg, PrecSea
# Save the Model as an object named DK.Biodiv.glm
DK.Biodiv.glmnb <- MASS::glm.nb(------- ~ ------- + ------- + ------- + ------- + -------, #  the Equations defining the response and additive combination of predictors
                          data = ------- # specify the Object where the data is in
                          )

# Call the summary() of the DK.Biodiv.glmnb object
-------(-------)
```
```{r glmNegBin-solution}
# Build a glm to predict S.AllGrp based on a linear combination of NatDensBasMp, Dis2Cost, HII, Slope30mAgg, PrecSea
# Save the Model as an object named DK.Biodiv.glm
DK.Biodiv.glmnb <- MASS::glm.nb(S.AllGrp ~ NatDensBasMp + Dis2Cost + HII + Slope30mAgg + PrecSea, #  the Equations defining the response and additive combination of predictors
                          data = DK.Biodiv # specify the Object where the data is in
                          )

# Call the summary() of the DK.Biodiv.glmnb object
summary(DK.Biodiv.glmnb)
```

<div class="alert alert-success">
**Question:** Is there an over-dispersion problem in this negative binomial model?

**WRITE YOUR ANSWER HERE**
</div>


<div class="alert alert-success">
**Question:** Which of the three models you build (`possion`, `quasipoisson`, `glm.nb`) you should use from now on?

**WRITE YOUR ANSWER HERE**
</div>

## Testing the asumptions.
<div class = "alert alert-info">
**Your task:**

Using the regression that addressed the over dispersion problem, check the assumption of a GLMM. For this yo will use the function `plot()` to visualize the diagnistic plot of the selected model.
</div>

```{r DIagPlot1, exercise=TRUE, exercise.lines=15, fig.dim=c(6,9)} 
# Set the plotting space to 2-rows and 2-columns
par(mfrow=c(2,2))
# Plot the object that contains the he regression that addressed the over dispersion problem
-------(-------)
```
```{r DIagPlot1-solution}
# Set the plotting space to 2-rows and 2-columns
par(mfrow=c(2,2))
# Plot the object that contains the he regression that addressed the over dispersion problem
plot(DK.Biodiv.glmnb)
```

<div class="alert alert-success">
**Question:** Are the corrected residuals (standardized residuals), normaly distributed?

**WRITE YOUR ANSWER HERE**
</div>


<div class="alert alert-success">
**Question:** Which plot do you need to look to assess if there is homogeneity of variances in this model? is there is homogeneity of variances>

**WRITE YOUR ANSWER HERE**
</div>


## How Good is my full model.	

<div class = "alert alert-info">
**Your task:**

For the Full-mode (`S.AllGrp ~ NatDensBasMp + Dis2Cost + HII + Slope30mAgg + PrecSea`), estimate the G^2^ statistic and its significance. For this, you will build a **CONSTANT** model (only intercept) and contrast it to `bolger.glm` using the `anova()` function.
</div>
  
```{r GSqrTst,  exercise=TRUE, exercise.lines=10, exercise=TRUE, exercise.lines=15} 
# Build a glm to predict RODENTSP based on a NO PREDICTOR save this as Const.glm
Const.glm.nb <- MASS::glm.nb(------- ~ -------, #  the Equations of a Constant model
                          data = ------- # specify the Object where the data is in
                          )

# Contrast the full model and the reduced model
-------(-------, # The FULL Model
        -------, # the Constant Model
        test = "Chisq") # What test to use
```
```{r GSqrTst-solution}
# Build a glm to predict RODENTSP based on a NO PREDICTOR save this as Const.glm
Const.glm.nb <- MASS::glm.nb(S.AllGrp ~ 1, #  the Equations of a Constant model
                          data = DK.Biodiv # specify the Object where the data is in
                          )

# Contrast the full model and the reduced model
anova(DK.Biodiv.glmnb, # The FULL Model
      Const.glm.nb, # the Constant Model
      test = "Chisq") # What test to use
```

<div class="alert alert-success">
**Question:** Is the **FULL** model better than a **CONSTANT** model?

**WRITE YOUR ANSWER HERE**
</div>  


## Which variables to keep 

<div class = "alert alert-info">
**Your task:**

Using the `step()` function and a "backward" section procedure you will build the Minimum Adequate Model to predict presence/absence of rodents.

</div>

  
```{r FinalModl1, exercise=TRUE, exercise.lines=15} 
# build the Minimum Adequate Model to predict presence/absence of rodents.
DK.Biodiv.glmnb.Back <- step(-------,
                              direction = "-------")
# Call bolger.glm.Back
DK.Biodiv.glmnb.Back
```
```{r FinalModl1-solution}
# build the Minimum Adequate Model to predict presence/absence of rodents.
DK.Biodiv.glmnb.Back <- step(DK.Biodiv.glmnb,
                              direction = "backward")
# Call bolger.glm.Back
DK.Biodiv.glmnb.Back
```
<div class = "alert alert-info">
**Your task:**

Using the `step()` function and a "forward" section procedure you will build the Minimum Adequate Model to predict presence/absence of rodents.
</div>

  
```{r FinalModl2, exercise=TRUE, exercise.lines=15} 
# build the Minimum Adequate Model to predict presence/absence of rodents.
DK.Biodiv.glmnb.Frwd <- step(-------,
                              direction = "-------")
# Call bolger.glm.Back
DK.Biodiv.glmnb.Frwd
```
```{r FinalModl2-solution}
# build the Minimum Adequate Model to predict presence/absence of rodents.
DK.Biodiv.glmnb.Frwd <- step(DK.Biodiv.glmnb,
                              direction = "forward")
# Call bolger.glm.Back
DK.Biodiv.glmnb.Frwd
```

<div class="alert alert-success">
**Question:** Do the two procedures select the same variables?

**WRITE YOUR ANSWER HERE**
</div>


## How good is my final model

<div class = "alert alert-info">
**Your task:**

Assess if your final-reduced model is better than a model with **NONE** of the predictors.</div>
  
```{r FinalModl3, exercise=TRUE, exercise.lines=15} 
# Contrast the full model and the reduced model
anova(-------, # The FULL Model
      -------, # the Constant Model
      test = "-------") # What test to use

## Based on the G2 test p-values, the Final Model is not different from a Full-model. Since we always prefer simple to complex models, the DK.Biodiv.Final model is preferred.
```
```{r FinalModl3-solution}
# Contrast the full model and the reduced model
anova(DK.Biodiv.glmnb.Back, # The FULL Model
      Const.glm.nb, # the Constant Model
      test = "Chisq") # What test to use

## Based on the G2 test p-values, the Final Model is not different from a Full-model. Since we always prefer simple to complex models, the DK.Biodiv.Final model is preferred.
```

<div class="alert alert-success">
**Question:** Is the reduced model better than a Constant model?

**WRITE YOUR ANSWER HERE**
</div>

## Testing the asumptions of the final model.
<div class = "alert alert-info">
**Your task:**

Check the assumption of a GLMM in your final model. For this, you will use the function `plot()` to visualize the diagnostic plot of the selected model.
</div>

```{r DIagPlot2, exercise=TRUE, exercise.lines=15, fig.dim=c(6,9)} 
# Set the plotting space to 2-rows and 2-columns
par(mfrow=c(2,2))
# Plot the object that contains the he regression that addressed the over dispersion problem
plot(-------)
```
```{r DIagPlot2-solution}
# Set the plotting space to 2-rows and 2-columns
par(mfrow=c(2,2))
# Plot the object that contains the he regression that addressed the over dispersion problem
plot(DK.Biodiv.glmnb.Back)
```
<div class="alert alert-success">
**Question:** Are the corrected residuals (standardized residuals), normaly distributed?

**WRITE YOUR ANSWER HERE**
</div>

<div class="alert alert-success">
**Question:** Is there homogeneity of variances?

**WRITE YOUR ANSWER HERE**
</div>