---
title: "Using `Spatial*` and `Raster` Data in R [Practical]"
author: '"YOUR NAME HERE"'
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
require(raster)
require(sp)
require(rgdal)
require(maptools)
require(rasterVis)
data(wrld_simpl)
```

# **The Set-up**

## *Learning objectives.*
1. Load, create, manipulate, and save `Spatial*` object using `R`.
2. Load, create, manipulate/extract, and save `Raster*` object using `R`.
3. Use map algebra to create and manipulate `Raster*` object using `R`.
4. Extract information from `Spatial*` and `Raster*`object using R.


## *Before you start*
The practical will be run using *RStudio cloud* to avoid individual machine troubleshooting. This means you **need** to have an account for *RStudio cloud*.

However, remember to have [`R`](https://cran.r-project.org) and [R-studio](http://www.rstudio.com/) in your/room computer so you can work on the project after class.

**You should have also **
1. Read chapters 2, 3, and 4 Bivand et al. Applied Spatial Data Analysis with R. Springer, 2013.
2. Done the Tutorial for this week.

## *The way it's going to be run*

In this 3h practical you will to implement what you have learned in the class about `Spatial*` and `Raster*`objects and how to Load, create, manipulate, and save these in `R`.

You will be building and **R-Markdown** document during the class. **R-Markdown** documents allows you to combine your code, its results, and your prose commentary into a single document. If you manage to produce a document your code works. The resulting file (a html) is what you will hand-in at the end of the practical.

**This is a 3h practical** and you will use all the time. Also try to write your own `R`-code and run it **DO NOT COPY THE CODE OF SOMEONE ELSE**. Writing your code and figuring out why it does not work is the best way to understand the dynamics of the `R` language and the tasks you are doing. Also, you have an excellent teaching assistant resource - ask him and/or me all the questions you have.

The final document will have for classes of boxes

<div class="alert alert-info">
<p class="text-center"; style = "font-size: 20px">
The text in the blue filled box marked as <strong>Your task</strong> states what you need to do</strong>
</p>
</div>

```{r }
#
# The code in the window marked like this shows where you need to write your code
#

```

<div class="alert alert-success">
<p class="text-center"; style = "font-size: 20px">
The text in the green filled box marked as <strong>Question</strong> phrases a question.
</p>
</div>

<div class="alert alert-success"; style = "background-color:White; color:black">
<p class="text-center"; style = "font-size: 20px">
Here you will place the answer to the <strong>question</strong>.
</p>
</div>


# **Obtaining environmental data for the Biowide dataset**

## *1. Load the Data*
<div class="alert alert-info">
**Your task:**

Load the data in *occurrence.csv*. The file is the folder associated to the project to the session - see the files tab in your R-studio session.. For this:

* Use `read.csv` to load the file -  The file is comma-separated 

* Save the object as `DK.Biodiv`.
</div>

```{r LoadData}
# Step 1. Read the file and save it as an object named `DK.Biodiv`
DK.Biodiv <-  
```

## *2. What is in the Table?*

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* Which groups (phylum) where sampled in the dataset?

</div>
```{r TableIn1}
# Step 1. Which groups where sampled in the dataset. Use the function unique() for this.


```

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* How many observations where recorded for each group (phylum)?

</div>
```{r TableIn2}
# Step 1. How many observations where recorded for each group (phylum). Use the function tapply() for this.


#
```

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* How many species where recorded for each group (phylum)?

* How many observations are in  the `DK.Biodiv` data set?

</div>
```{r TableIn3}
#Step 1. How many species where recorded for each group (phylum). Use the function tapply() for this.


#Step 2. How many observations where recorded for each group (phylum). Use the function tapply() for this.

```

<div class="alert alert-success">
**Question:**

* What is the Difference between the number of observations and the number of species per groups?

</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, estimate the avenge number of species per sampled site of the groups *Magnoliophyta* and *Pinophyta*. For this, you will:

* Add a new column to `DK.Biodiv` named *siteID* that define the unique locations by combining the `Lat` & `Lon` data. Use the function `paste()` or `paste0()` for this.

* Count the number of species per site for *Magnoliophyta*. Store the result in an object named `Rich.Magnoliophyta`.

* Count the number of species per site for *Pinophyta*. Store the result in an object named `Rich.Pinophyta`.

* Estimate the average number of species per site using the data in `Rich.Magnoliophyta`  `Rich.Pinophyta`.

</div>
```{r TableIn4}
# Avenge number of species per sampled site of the groups *Magnoliophyta* and *Pinophyta*?

# Step-1 Define the unique Locations based on Lat-Lon data - add it as a column to the data.frame
DK.Biodiv$siteID <- 

# Step 2 Count the number of species per site. Use the function tapply() for this.
Rich.Magnoliophyta <- 
  
# Step 3 Count the number of species per site. Use the function tapply() for this.
Rich.Pinophyta <- 
  
# Step 4 Estimate the average No of species Per site
# Estimate the Average for Magnoliophyta

# Estimate the Average for Pinophyta
```

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will:

* Estimate the 95%-confidence interval in the number of species per sampled site of the groups *Magnoliophyta* and *Pinophyta*  using the `quantile()` function.

</div>

```{r TableIn5a}
# 95%-confidence interval in the  number of species per sampled site of the groups *Magnoliophyta* and *Pinophyta*?

# Step 1 Estimate the 95%-confidence interval No of species Per site - using quantile
# Estimate the quantile for Magnoliophyta

# Estimate the quantile for Pinophyta

```
<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will:

* Estimate the 95%-confidence interval in the number of species per sampled site of the groups *Magnoliophyta* and *Pinophyta*  using a using a normal approximation.

</div>
```{r TableIn5b}
# 95%-confidence interval in the  number of species per sampled site of the groups *Magnoliophyta* and *Pinophyta*?

# Step 1 Estimate the 95%-confidence interval No of species Per site - using a normal approximation
## 95%-CI for Magnoliophyta

## 95%-CI for Pinophyta

```

<div class="alert alert-success">
**Question:**

* What is the Difference between using the `quantile` function and a normal approximation to estimate 95%-Confidence intervals?

</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* Count the number of species per site per group using the function `tapply()`.

* Plot the box-plot of the **Log_10* of the number of species per site sites for all evaluated groups.

</div>

```{r TableIn6}
# box-plot of the **Log_10* of species richness.

# Step 1 Count the number of species per site per group. Use the function tapply() for this.
Rich.PerSite <- 
  
# Step 2. Plot the Box-plot of log-10 richness

```


<div class="alert alert-success">
**Question:**

* How do you interpret a difference in one (1) unit in Y-axis of the box-plot you just plotted?

</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will:

* Transform `Rich.PerSite` to from a matrix into a `data.frame` using the function `as.data.frame()`.

* Define the latitude to each site, and add this value to `Rich.PerSite` as a new variable named *Lat*.

* Define the longitude to each site, and add * Define the latitude to each site, and add this value to `Rich.PerSite` as a new variable named *Lat*. value to `Rich.PerSite` as a new variable named *Lon*.

* Estimate the correlation between species richness per site of vascular plants (*Magnoliophyta* and *Pinophyta*) and latitude.

* Estimate the correlation between species richness per site of vascular plants (*Magnoliophyta* and *Pinophyta*) and longitude.

</div>

```{r TableIn7}
# Estimate the correlation between species richness per site and latitude.

# Step 1 transform Rich.PerSite to a data.frame
Rich.PerSite <- 

# Step 2 Define the latitude to each site. Use the function strsplit() for this.
Rich.PerSite$Lat <-

# Step 3 Define the Longitude to each site. Use the function strsplit() for this.
Rich.PerSite$Lon <- 

# Step 4 estimate the correlation between richness and Latitude. Use the function apply() for this.

# Step 5 estimate the correlation between richness and longitude. Use the function apply() for this.

```

## *3. Creating and working with `SpatialPointsDataFrame`s*

<div class="alert alert-info">
**Your task:**
  
Based on the data stored in the object `Rich.PerSite`, you will create a `SpatialPointsDataFrame` were each point represents a sampled location. For this you will:

* Copy the data in `Rich.PerSite` into a new object named `BiowideSptPnt`.

* Use the function `coordinates()` to transform `BiowideSptPnt` into a `SpatialPointsDataFrame`.

* Check the class of `BiowideSptPnt`.

* Call the `BiowideSptPnt` object.

</div>

```{r SpatPnt1}
# Create a `SpatialPointsDataFrame` 
BiowideSptPnt <- 

#Step 1. Use the function coordinates() to create a SpatialPointsDataFrame.

#Step 2. Check the class of BiowideSptPnt to be sure that is a SpatialPointsDataFrame. 


#Step 2. Call the BiowideSptPnt object

```
<div class="alert alert-info">
**Your task:**
  
Using the object `BiowideSptPnt`, you will.

* Check the projection of `BiowideSptPnt`.

* Define a Projection for `BiowideSptPnt` using the function `projection()`.

* Visualize the `BiowideSptPnt` object.
</div>

```{r SpatPnt2}
# Define a Projection for BiowideSptPnt

#Step 1. Check the projection of BiowideSptPnt

# Step 2. Define the projection for BiowideSptPnt
## As the Object has no projection you can just add it as a string. The string for WGS84 is "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs". 

# Step 3. Visualize the BiowideSptPnt object

```

<div class="alert alert-info">
**Your task:**
  
Based on the data stored in the object `Rich.PerSite`, create a `SpatialPointsDataFrame` were each point represents a sampled location. For this, you will:

* Use the function `SpatialPointsDataFrame()` to transform `Rich.PerSite` into a spatial object named `BiowideSptPnt`.

* Check the class of `BiowideSptPnt`.

* Call and visualize the `BiowideSptPnt` object.
</div>

```{r SpatPnt3}
# Step 1. Create a `SpatialPointsDataFrame` 
BiowideSptPnt <- 

# Step 2. Check the class of BiowideSptPnt to be sure that is a SpatialPointsDataFrame. 

# Step 3a. Call the BiowideSptPnt object

# Step 3b. Visualize the BiowideSptPnt object

```

<div class="alert alert-success">
**Question:**

* What is the diff between the the `SpatialPointsDataFrame` created with the function `coordinates()` and the one created using the `SpatialPointsDataFrame()` function?

</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**
  
Based on the data stored in the object `BiowideSptPnt`, you will.

* Re-project this object to the projection system for Denmark (UTM/ETRS89).

* Call and visualize `BiowideSptPnt.UTM`.
</div>

```{r SpatPnt4}
# Step 1. Re-project BiowideSptPnt to UTM/ETRS89
BiowideSptPnt.UTM <- 
  
# Step 2a. Call BiowideSptPnt.UTM


# Step 2b. plot BiowideSptPnt.UTM

```

## *4. Creating and working with `SpatialPolygonsDataFrame`s*

<div class="alert alert-info">
**Your task:**
Here you will get data from an on-line repository. For this. you will:

* Use the function `getData()` to get the first level of administrative subdivision for Denmark (i.e., landsdele). Store the data in an object named `LandsdeleShp`.

* Visualize `LandsdeleShp`.
</div>

```{r SpaPoly1}
# Step 1. Get The level-1 map for Denmark - Denmark/ISO-code is DNK. Use the function getData() for this. Also, Denmark ISO code is DNK.
LandsdeleShp <- 

  # Step 2. Visualize LandsdeleShp

```

<div class="alert alert-success">
**Question:**

* Can you use the object `BiowideSptPnt.UTM` to extract the names of the regions of Denmark in `LandsdeleShp` for each sampled location?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**
Based on the data stored in the object `LandsdeleShp`, you will.

* Call the projection of `LandsdeleShp`.

* Re-project this object to the projection system for Denmark (UTM/ETRS89). For this, use the function `spTransform()`.

* Call and plot `LandsdeleShp.UTM`.

</div>
```{r SpaPoly2}
# Step 1. check the projection of LandsdeleShp


# Step 2. Re-project LandsdeleShp to UTM/ETRS89
LandsdeleShp.UTM <- 
  
# Step 3a. Call LandsdeleShp.UTM to check the projection of LandsdeleShp.UTM


# Step 3b. Plot LandsdeleShp.UTM

```

<div class="alert alert-info">
**Your task:**
Using the `BiowideSptPnt.UTM` and `LandsdeleShp.UTM` objects, you will.

* Call the data included in `LandsdeleShp.UTM`.

* Extract the region of Denmark (Landsdele) where each variable was sampled. For this you will use the function `over()`.

* Add this information to `BiowideSptPnt.UTM` as a new variable named *landsdele*.

* Check the data in `BiowideSptPnt.UTM` by calling the first five rows.

</div>
```{r SpaPoly3}
# Step 1. Call the data included in LandsdeleShp.UTM


# Step 2. Extract the region name that corresponds to each sampled point using the function over. Use the function over() for this.
landsdele <- 

# Step 3. add this variable to BiowideSptPnt.UTM
BiowideSptPnt.UTM$landsdele <- 

# Step 3. Check the data in BiowideSptPnt.UTM by Calling the first five rows

```
<div class="alert alert-info">
**Your task:**
Using the `BiowideSptPnt.UTM` and `LandsdeleShp.UTM` objects, you will.

* Estimate the number of sampled sites per region and save this summary as an object named `SmpSitPerReg`.

* Add the values in `SmpSitPerReg` to`LandsdeleShp.UTM` a new variable that corresponds to the number of sampled sites per region. Name this new variable *NoSampSites*.

* Call the data included in `LandsdeleShp.UTM`.

</div>
```{r SpaPoly4}
# Step 1. Estimate the number of sampled sites per region. Use the function table() for this.
SmpSitPerReg <- 

# Step 2b. add this variable to LandsdeleShp.UTM
LandsdeleShp.UTM$NoSampSites <- 

# Step 3. Check the data in LandsdeleShp.UTM

```
<div class="alert alert-info">
**Your task:**
Using the `BiowideSptPnt.UTM` and `LandsdeleShp.UTM` objects, you will.

* Plot`LandsdeleShp.UTM` colouring the Regions (landsdele) by the number of sampling sites.

* Add a legend can be used to interpret the colours representing the number of sampled sites. Use the function `legend()` for this.

* Estimate the richness of vascular plants per site by adding the richness of *Magnoliophyta* & *Pinophyta*.

* Overlay on top of this map `BiowideSptPnt.UTM` sizing the points by the richness of vascular plants.

* Add a legend that can be used to interpret the species richness per site.

</div>

```{SpaPoly4}
# Step 1. Plot`LandsdeleShp.UTM` colouring the Regions

# Step 2. Add a legend can be used to interpret the colours representing the number of sampled.  

# Step 3. Estimate the richness of vascular plants. Use the function apply() for this.
S.Spermatophyte <- 
  
# Step 4. Overlay `BiowideSptPnt.UTM`. Used the argument add=T for this.

# Step 5. Add a legend that can be used to interpret the species richness per site.  

```

## *5. Rasters-1: Reclassify and aggregate a 'raster' object*

<div class="alert alert-info">
**Your task:**
Load the data in *BASEMAP_2018_Crop.tif*, the file is the folder associated to the project to the session - see the files tab in your R-studio session.

* Use `raster()` to load the file.

* Save the object as `BASEMAP.Crop`.

* Visualize the object.
</div>
```{r RasrRecl1}
# Step 1. Load BASEMAP_2018_Crop.tif
BASEMAP.Crop <- 
# Step 2. Call the BASEMAP.Crop object

# Step 3. visualize the BASEMAP.Crop object

```

<div class="alert alert-success">
**Question:**

* What is the projection of `BASEMAP.Crop`?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**
Using the object `BASEMAP.Crop` you will:

* Load the *Basemap-Legend.csv* file with the link between codes in `BASEMAP.Crop`and the land cover classes names.The file is in the folder associated to the project to the session.

* Extract the Data in `BASEMAP.Crop` and save it as a object named `BASEMAP.Data`.

* Transform `BASEMAP.Data` to a factor using the data in `BASEMAP.Legend`.

* Estimate the number of cells per Land cover class. Use the function `table()` to do this.
</div>
```{r RasrRecl2}
# Step 1. Load Basemap-Legend.csv
BASEMAP.Legend <- 

# Step 2. Extract the Data in BASEMAP.Crop and save it as a object named BASEMAP.Data
BASEMAP.Data <- 

# Step 3. Transform BASEMAP.Data to a factor using the data in BASEMAP.Legend to define the levels and labels. Use the function factors() for this.
BASEMAP.Data.Factor <- 
  
# Step 4. Use the function table to count the number of cells per land cover type defined in BASEMAP.Data.Factor. Use the function table() for this.
CountLandCov <- 
  
# Call CountLandCov.

# Step 5. Sort the data in CountLandCov in increasing order

```

<div class="alert alert-success">
**Questions:**

* Which land Cover type is the most abundant?

* Which land Cover type(s) are not represented?

* What is the coverage of Forested areas?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**
Using the object `BASEMAP.Crop` and the data in `BASEMAP.Legend` you will reclassify ``BASEMAP.Crop` so that it shows areas consider as Agricultural areas (Agriculture-intensive/extensive-temporary crops/permanent crops). For this, you will:

* Create a new table that transform the *Names* variable in `BASEMAP.Legend` to 1 (Agricultural areas) and 0 (Other areas). Name this new table `Reclas.Table`.

* Transform `Reclas.Table` into a matrix. Use the function `as.matrix()`.

* Reclassify `BASEMAP.Crop` using `Reclas.Table` so that only only shows the cells considered as Agricultural areas. For this, use the function `reclassify()`. Store the reclassified map as an object named `BASEMAP.Agri`.

* Plot `BASEMAP.Agri`.

</div>
```{r RasrRecl3}
# Step 1a. Create a reclassification table (Reclas.Table) using the data in BASEMAP.Legend
Reclas.Table <- 
# Step 1b. make the column Name in Reclas.Table all Zeros (0)
Reclas.Table$Name <- 0
# Step 1c. Change the value in Reclas.Table$Name for all Agricultural sites 1

# Step 1d. Add a row to the end of Reclas.Table that makes all NA values NA. Use the function rbind() for this.

# Step 2. Transform Reclas.Table into a matrix
Reclas.Table <- 

  # Step 3. Use the reclassify function to transform BASEMAP.Crop into a binary map (BASEMAP.Agri) based on Reclas.Table. Use the function reclassify() for this
BASEMAP.Agri <- 
  
# Step 3. plot BASEMAP.Agri

```

<div class="alert alert-info">
**Your task:**
Using the object `BASEMAP.Agri` you will:

* Estimate the Agricultural Density per 1km^2 using the function `aggregate()`. save the aggregated raster as an object named `BASEMAP.Agri`.

* Plot `BASEMAP.Agri`.
</div>
```{r RasrRecl4}
# Step 1. Aggregate the raster `BASEMAP.Agri`. Save the result as an object named AgriDen. Use the function aggregate() to make this summary.
AgriDen <- 
  
# Step 2. Plot the AgriDen object to visualize the result

```

<div class="alert alert-info">
**Your task:**
Load the data in *NatDens_BaseMap.tif* that has the density of natural areas based on BASEMAP. The file is the folder associated to the project to the session - see the files tab in your R-studio session.

* Use `raster()` to load the file.

* Save the object as `NatDens.DK`.

* Call the object to check it's properties.

* Visualize the object.
</div>
```{r RasrRecl5}
# Step 1. Load NatDens_BaseMap.tif
NatDens.DK <- 

# Step 2. Call the NatDens.DK object


# Step 3. Visualize the NatDens.DK object

```

<div class="alert alert-success">
**Questions:**

* What is the projection of the `NatDens.DK` raster?

* Can you use the object `BiowideSptPnt` to extract the values of Nature Density from `NatDens.DK`.
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**
Extract the information in `NatDens.DK` using the appropriate Biowide
`StaialPointsDataFrame` object. For this, you will:

* Use the function `extract ()`.

* Save the data as a new variable named *NatDens* in the appropriate Biowide
`StaialPointsDataFrame` object.

* Plot a histogram for *NatDens*. 
</div>

```{r RasrRecl6}
#Step 1. Extract the data in NatDens.DK & save it to the appropriate SpatialPoints object. Use the function extract() for this. 
  
#Step 2. Plot a histogram for *NatDens* in BiowideSptPnt.UTM

```

<div class="alert alert-info">
**Your task:**
Based on the information in `BASEMAP.Crop` and `BASEMAP.Legend` you will plot the land cover raster as a categorical raster. For this, you will:

* Transform `BASEMAP.Crop` into a factor using the function `ratify()`. Store the transformed `BASEMAP.Crop` as an object names `BASEMAP.Crop.Fact`.

* Extract the levels in the `BASEMAP.Crop.Fact` categorical raster using the function `levels()`. Store these levels as an object named `rat`.  

* Add an extra attribute to `rat` that will be used for the legend. This is done by matching the values in `rat` to the `code` variable in `BASEMAP.Legend`, and extracting  from `BASEMAP.Legend` the variable `Name`. use the function `match()` for this.

* Make now `rat` the `leves` of `BASEMAP.Crop.Fact` using the function `levels()`.

* Plot  `BASEMAP.Crop.Fact` using the function `levelplot()` from the `rasterVis` package.

</div>

```{r RasrRecl7}
# Step 1. Transform `NatDens.DK` into a factor. Use the function ratify() for this.
BASEMAP.Crop.Fact <- 

# Step 2. Extract the Factors - note that you will get a list with one slot. Use the function levels() for this.
rat <- 

# Step 3. Add the Name for each factors using the information in `BASEMAP.Legend`
rat[[1]][,'landcover'] <- 

# Step 4. Make rat the levels in BASEMAP.Crop.Fact. Use the function levels() for this.

# Step 4. Plot BASEMAP.Crop.Fact. Use the function levelplot() for this. Set the colors using the function rainbow().

```

## *5. Rasters-3: Map Algebra*

<div class="alert alert-info">
**Your task:**
Load the data in *Tmean_CHELSA-DK.tif*, the file is the folder associated to the project to the session - see the files tab in your R-studio session.

* Use `stack()` to load the file.

* Save the object as `Tmean.DK`.

* Visualize the object.
</div>

```{r RasrMapAlg1}
# Step 1. Load Tmean_CHELSA-DK.tif and store it as an object named Tmean.DK
Tmean.DK <- 

# Step 2. Visualize the object.

```
<div class="alert alert-success">
**Question:**

* What is the difference between loading the data using the functions `raster()`, `stack()`, and `brick()`
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**
Using the data in `Tmean.DK` yo will now estimate the annual Temperature seasonality as:

* The standard deviation of mean monthly temperatures. Store this operation as an object named `TSes.SD`. Use the function `calc()` for this.

* The coefficient of variation of mean monthly temperatures. Store this operation as an object named `TSes.CV`. Use the function `calc()` for this.

* Stack `TSes.SD` and `TSes.CV` as an object named `TSes.DK`. Use the function `stack()` for this.

* Visualize `TSes.DK`.
</div>

```{r RasrMapAlg2}
# Step 1. Estimate the annual Temperature seasonality as SD. Use the function calc() for this.
TSes.SD <- 

# Step 2. Estimate the annual Temperature seasonality as CV. Use the function calc() for this.
TSes.CV <- 

# Step 3. Stack `TSes.SD` and `TSes.CV`. Use the function stack() for this.
TSes.DK <- 
names(TSes.DK)<- c("TSes.SD","TSes.CV")

# Step 4. Visualize `TSes`.

```

<div class="alert alert-success">
**Questions:**

* What is the projection of the `TSes.DK` raster?

* Can you use the object `BiowideSptPnt.UTM` to extract the values of Nature Density from `NatDens.DK`.
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**PLACE YOUR ANSWER HERE**
</div>

<div class="alert alert-info">
**Your task:**
Using the object `TSes.DK` you will:

* Re-project the raster to  Denmark's projection system (UTM/ETRS89). For this, use the function `projectRaster()`.

* Store the Re-projected raster as an object named `TSes.DK.UTM` 

* Call the `TSes.DK.UTM` to check the projection.
</div>

```{r RasrMapAlg3}
# Step 1. Re-project the raster. Use the function projectRaster() for this.
TSes.DK.UTM <- 

# Step 2. Call the Re-projected raster

```

<div class="alert alert-info">
**Your task:**
Using the objects `TSes.DK` and `BiowideSptPnt.UTM` you will:

* Extract both the Temperature seasonality estimated using the standard deviation for all sampled locations.

* Save the data as a new variable named *TSes* in `BiowideSptPnt.UTM`.

* Visualize the extracted data as a box-plot grouping *TSes* by *regions*.
</div>

```{r RasrMapAlg4}
# Step 1. Extract both the SD derived Temperature seasonality. Use the function extract() for this.
BiowideSptPnt.UTM$TSes <- 

# Step 2. Visualize TSes as a box-plot

```


