---
title: "DataExploration_Practical"
author: '"YOUR NAME HERE"'
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **The Set-up**

## *Learning objectives.*
1.	Load tables into `R` and explore their content.
2.  Execute basic looping routines using the `*apply()` and `for()` functions.
3.  Graphically explore the variability in the variables in a data set based.  
4.	Graphically compare the variables values between categorical groups  
5.	Establish the normality of a variable and evaluate different transformations to achieve normality.
6.	Determine the correlation between two random variables and the significance of such correlation.  


## *Before you start*
The practical will be run using *RStudio cloud* to avoid individual machine troubleshooting. This means you **need** to have an account for *RStudio cloud*.

However, remember to have [`R`](https://cran.r-project.org) and [R-studio](http://www.rstudio.com/) in your/room computer so you can work on the project after class.

**You should have also **
1. Read chapter 2 of Borcard et al. Numerical ecology with R. Springer, 2018.
2. Done the Tutorial for this week.

## *The way it's going to be run*

In this 3h practical you will to implement what you have learned in the class about Exploratory Data Analysis in `R`.

You will be building and **R-Markdown** document during the class. **R-Markdown** documents allows you to combine your code, its results, and your prose commentary into a single document. If you manage to produce a document your code works. The resulting file (a html) is what you will hand-in at the end of the practical.

This 3h practical is divided into two parts. **DO NOT COPY AND PASTE THE CODE!** 

Also try to write your own `R`-code and run it **DO NOT COPY THE CODE OF SOMEONE ELSE**. Writing your code and figuring out why it does not work is the best way to understand the dynamics of the `R` language and the tasks you are doing. Also, you have an excellent teaching assistant resource - ask him and/or me all the questions you have.

The final document will have for classes of boxes

<div class="alert alert-info">
<p class="text-center"; style = "font-size: 20px">
<strong>
The text in the blue filled box marked as **Your task** states what you need to do</strong>
</strong>
</p>
</div>

```{r }
#
# The code in the window marked like this shows where you need to write your code
#

```

<div class="alert alert-success">
<p class="text-center"; style = "font-size: 20px">
<strong>The text in the green filled box marked as **Question** phrases a question.
</strong>
</p>
</div>

<div class="alert alert-success"; style = "background-color:White; color:black">
<p class="text-center"; style = "font-size: 20px">
<strong> Here you will place the answer to the **question**.
</strong>
</p>
</div>


# **Exploring Denmark's biodiversity and environmental data.**

##*1. Load the Data*
<div class="alert alert-info">
**Your task:**

Load the data in *Biowide_With_EnvDta.csv*, located in the Web URL <https://raw.githubusercontent.com/AlejoOrdonez/StaGeoMod2022/main/Week%201%20-%20Data%20Exploration/Practical/Data/Biowide_With_EnvDta.csv>. For this:

* Define an object named URL.Loc with the string of the Web URL.

* Use `read.csv` to load the file -  The file is comma-separated 

* Save the object as `DK.Biodiv`.
</div>

```{r LoadData}
# Define an object named URL.Loc with the string of the Web URL.

URL.Loc <- "https://raw.githubusercontent.com/AlejoOrdonez/StaGeoMod2022/main/Week%201%20-%20Data%20Exploration/Practical/Data/Biowide_With_EnvDta.csv"

# Read the file and save it as an object named `DK.Biodiv`
DK.Biodiv <-  read.csv(URL.Loc)

#
```

##*2. What is in the Table?*

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* What is the object class of `DK.Biodiv` data set?

</div>
```{r TableIn1}
# What is the object class of `DK.Biodiv` data set?
class(DK.Biodiv)

#
```

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* Which are the variable names in the `DK.Biodiv` data set?

</div>
```{r TableIn2}
#Which are the variable names in the `DK.Biodiv` data set?
names(DK.Biodiv)

#
```

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* How many observations are in  the `DK.Biodiv` data set?

</div>
```{r TableIn3}
#How many observations are in  the `DK.Biodiv` data set?
dim(DK.Biodiv)[1]

#
```

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* How many variables are in  the `DK.Biodiv` data set?

</div>
```{r TableIn4}
#How many variables are in  the `DK.Biodiv` data set?
dim(DK.Biodiv)[2]
#or
length(names(DK.Biodiv))

#
```

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* List the variables that are `numeric`, and `character`.

</div>
```{r TableIn5}
#List the variables that are `numeric` and `character`.

# Numeric variables
NumVar <- names(DK.Biodiv)[(sapply(DK.Biodiv,class)%in%c("numeric","integer"))]
NumVar

# Character variables
CharVar <- names(DK.Biodiv)[(sapply(DK.Biodiv,class)%in%c("character"))]
CharVar

```

<div class="alert alert-success">
**Question:** What is the Difference between `numeric` and `integer` objects?

</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">

**Numeric** data consists of numbers with decimals points. **Integer** data consists of whole numbers.

Operational they are the same, but in `R` **integers** are subsets of **numeric**.
</div>

##*3. Summary statistics*

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will estimate.

* The mean of all `numerical` variables. You **MUST** use an `apply()` function in this task.
</div>
```{r SummStat1}
# mean of all `numerical` variables.
Mean.DK.Biodiv <- apply(X = DK.Biodiv[,NumVar], # Data frame with the Data
                        MARGIN = 2, # Dimension on which the function will be applied over
                        FUN = mean, # Function to be applied over
                        na.rm = T # Do not consider NA values
                        )
Mean.DK.Biodiv
#
```

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will estimate.

* The standard deviation of all `numerical` variables. You **MUST** use a `sapply()` or `lapply()` function in this task.
</div>
```{r SummStat2}
# The standard deviation of all `numerical` variables.
SD.DK.Biodiv <- sapply(X = NumVar, # a vector defining the looping space
                       function(x){ # Function to be applied over
                         sd(x = DK.Biodiv[,x], # a numeric vector 
                            na.rm = T # Do not consider NA values
                            )
                         })

SD.DK.Biodiv
#
```

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will estimate.

* The [coefficient of variation](https://en.wikipedia.org/wiki/Coefficient_of_variation) for all `numerical` variables.
</div>
```{r SummStat3}
# The coefficient of variation for all `numerical` variables.
CV.DK.Biodiv <- Mean.DK.Biodiv/SD.DK.Biodiv
CV.DK.Biodiv
# OR
CV.DK.Biodiv <- apply(X = DK.Biodiv[,NumVar], # Data frame with the Data
                        MARGIN = 2, # Dimension on which the function will be applied over
                        FUN = function(x){mean(x,na.rm = T)/sd(x,na.rm = T)} # Function to be applied over
                        )
CV.DK.Biodiv
#
```

<div class="alert alert-success">
**Question:** Can you compare the coefficient of variations of different variables? is yes/no, explain why?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**Yes** you can compare the coefficient of variations (CV) of different variables as these are are measurements of relative variability in data sets.

The CV scales the variability measured by the standard deviation by the mean, so it show proportional changes which are comparable between groups.
</div>


<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will estimate.

* The mean of all `numerical` variables per region. You **MUST** use a combination of `apply()` and `tapply()` function in this task. 
</div>
```{r SummStat4}
# The mean of all `numerical` variables per region. 
RegMean.DK.Biodiv <-  apply(X = DK.Biodiv[,NumVar], # Data frame with the Data
                        MARGIN = 2, # Dimension on which the function will be applied over
                        FUN = function(x){# Function to be applied over
                          tapply(X = x, # vector to aggregate
                                 INDEX = DK.Biodiv$landsdele, # The grouping factor
                                 FUN = mean, # Function to be applied over
                                 na.rm = T # Do not consider NA values
                                 )
                        })
RegMean.DK.Biodiv

```

##*3. Graphical representations*

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will write code to plot:

*  A bar-plot that summarises the mean *Nature density* (`NatDensBasMp`) across observations in each region (`landsdele`).

*Add to this bar plot the [95% CI error bars](https://en.wikipedia.org/wiki/Confidence_interval) using the function `arrows()`.

* You **MUST** use a `tapply()` function to generate the necessary summaries (i.e., Means, SD, and Number of observations).

</div>

```{r Graphical1}
#A bar-plot that summarises the mean *Nature density* (`NatDensBasMp`) across observations in each region (`landsdele`).
# Add to this bar plot the 95% CI error bars using the function `arrows()`

## Estimate the mean Nature density per region
NatDen.Mean <- tapply(X = DK.Biodiv$NatDensBasMp, # Data frame with the Data
                      INDEX = DK.Biodiv$landsdele, # The grouping factor
                      FUN = mean, # Function to be applied over
                      na.rm = T # Do not consider NA values
                      )
NatDen.Mean

## Estimate the Nature density sd per region
NatDen.SD <- tapply(X = DK.Biodiv$NatDensBasMp, # Data frame with the Data
                      INDEX = DK.Biodiv$landsdele, # The grouping factor
                      FUN = sd, # Function to be applied over
                      na.rm = T # Do not consider NA values
                      )
NatDen.SD

## Estimate the number of observation  per region                      
NatDen.n <- table(DK.Biodiv$landsdele)
NatDen.n

## Determine the Low an high value of the 95% CI
LowCI <- NatDen.Mean - qnorm(0.976)*(NatDen.SD/sqrt(NatDen.n))
HighCI <- NatDen.Mean + qnorm(0.976)*(NatDen.SD/sqrt(NatDen.n))

## Plot the barplot
BarsNat <- barplot(height = NatDen.Mean, # Variable to plot
                   main = "Nature density", # Add the variable name as title
                   ylim = c(0,max(HighCI)) # Determine the range of the y-axis
                   )

## Plot the Arrows with the 96%CI
arrows(x0 = BarsNat, #coordinates of points from which to draw
       y0 = LowCI, #coordinates of points from which to draw
       y1 = HighCI, #coordinates of points from which to draw
       angle = 90, # angle of the arrow (90 is flat)
       length = 0.05, # length of the edges of the arrow head 
       code = 3, # Arrowhead in on or both ends
       )
#
```

<div class="alert alert-success">
**Question:** DO you see differences in *Nature density* of the sampled sites between regions? Explain why yes/no.
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
Yes, there are differences between **GROUP** of regions.

On group is composed by Hovedstaden, Midtjylland, and Nordjylland; and a higher Nature density.

The second group is composed by Sjælland and Syddanmark, and has comparative lower Nature density.
</div>

<div class="alert alert-info">

**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will write code to plot:

*  A bar-plot that summarises the mean and error bars of *Human Impact* (`HII`) per [quantile](https://en.wikipedia.org/wiki/Quantile) of *total species richness* (`S.AllGrp`).

* You **MUST** use the `quantile()` and `findInterval()` functions to determine tow wich *total species richness* quantile each observation belongs.

* You **MUST** use a `tapply()` function to generate the necessary summaries (i.e., Means, SD, and Number of observations).

</div>

```{r Graphical2}
# A bar-plot that summarises the mean and error bars of *Human Impact* (`HII`) per quantile of species richness (`S.AllGrp`). 

## Estimate the quantiles for S.AllGrp 
RichQuant <- quantile(x = DK.Biodiv$S.AllGrp, # Vector to quantify
                      na.rm = T # Do not consider NA values
                      )

# Determine to which quantile each observation belongs
RichQuantSumm <- findInterval(x = DK.Biodiv$S.AllGrp, # Vector to classify
                              vec = RichQuant, # A vector with the intervals
                              rightmost.closed = T, # Determine if the max value correspond to clas N or N-1
                              )

## Estimate the mean Human Impact per quantile
HII.Mean <- tapply(X = DK.Biodiv$HII, # Data frame with the Data
                   INDEX = RichQuantSumm, # The grouping factor
                   FUN = mean, # Function to be applied over
                   na.rm = T # Do not consider NA values
                   )
HII.Mean

## Estimate Human Impact sd per quantile
HII.SD <- tapply(X = DK.Biodiv$HII, # Data frame with the Data
                      INDEX = RichQuantSumm, # The grouping factor
                      FUN = sd, # Function to be applied over
                      na.rm = T # Do not consider NA values
                      )
HII.SD

## Estimate the number of observation  per quantile                      
HII.n <- table(RichQuantSumm)
HII.n

## Determine the Low an high value of the 95% CI
LowCI <- HII.Mean - qnorm(0.976)*(HII.SD/sqrt(HII.n))
HighCI <- HII.Mean + qnorm(0.976)*(HII.SD/sqrt(HII.n))

## Plot the barplot
BarsHII <- barplot(height = HII.Mean, # Variable to plot
                   main = "Nature density", # Add the variable name as title
                   ylim = c(0,max(HighCI)), # Determine the range of the y-axis
                   names.arg = paste(1:4,"- quantile") # Names under each bar
                     )

## Plot the Arrows with the 96%CI
arrows(x0 = BarsHII, #coordinates of points from which to draw
       y0 = LowCI, #coordinates of points from which to draw
       y1 = HighCI, #coordinates of points from which to draw
       angle = 90, # angle of the arrow (90 is flat)
       length = 0.05, # length of the edges of the arrow head 
       code = 3, # Arrowhead in on or both ends
       )
#
```

<div class="alert alert-success">
**Question:** Do you see differences in *human Impact* between quantiles of species richness? What is the implication of the existence or not of diferecnes?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
There are not differences in *human impact* between quantiles of species richness.

This pattern indicates no relation between *human impact* and changes in species richness.
</div>

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will write code to plot:

* A box-plot that summarises the variation in *Slopes* (`Slope30mAgg`) across observations in each region (`landsdele`). In this figure you will colour code each box-plot using a [hcl diverging pallets](https://blog.r-project.org/2019/04/01/hcl-based-color-palettes-in-grdevices/index.html).
</div>

```{r Graphical3}
# A box-plot that summarises the variation in *Slopes* (`Slope30mAgg`) across observations in each region (`landsdele`).
# colour code each box-plot using a hcl diverging pallet.


## Set the colour per Region
ClPerReg <- hcl.colors(n = length(unique(DK.Biodiv$landsdele)),# the number of colours
                       palette = 'RdYlBu' # palette name, Here a red to yellow to blue
                       ) 
## Draw the Boxplot

boxplot(Slope30mAgg ~ landsdele, # formula defining the variable to summarize and the grouping
        data = DK.Biodiv, # the Data frame with the Data
        col = ClPerReg # Thw color of each box.plot
        )
```

<div class="alert alert-success">
**Question:** What is the box plot of *SLopes* per region show?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
The box-plot shows that wile the median level of variation (the horizontal thick line) is the same accords regions, samples in Midtjylland and Sjælland have a broader range of variability (1st and 2nd quantiles shown in the boxes) and extremes (the whiskers).
</div>

##*4. Normality tests*

For the rest of the practical you will use a subset of the data stored in the object `DK.Biodiv`. The subset will include the flowing variables:

1) *Richness of all sampled species* [`S.AllGrp`]
2) *Region* [`landsdele`]
3) *Parental material* [`Geology`]
4) *Nature Density* [`NatDensBasMp`]
5) *Distance to coast* [`Dis2Cost`]
6) *Human Impact Index* [`HII`]
7) *Slope* [`Slope30mAgg`]
8) *Precipitation Seasonality* [`PrecSea`]

To make things easy you will now bundle these into a new object called 

<div class="alert alert-info">
**Your task:**

Based on the data stored in the `DK.Biodiv` object, you will write code to subset the main data object into a `data.frame` named `DK.Biodiv.Smpl` that **ONLY** includes the eigth key variables to evaluate.
 
</div>

```{r Normality1}
## Variable to use
VarUse <- c("S.AllGrp", "landsdele", "Geology", "NatDensBasMp", "Dis2Cost", "HII", "Slope30mAgg", "PrecSea")

# Sub sample the the `DK.Biodiv` object
DK.Biodiv.Smpl <- DK.Biodiv[,VarUse]
  
# check the new data.frame  
head(DK.Biodiv.Smpl)
```

<div class="alert alert-info">
**Your task:**

* Based on the data stored in the `DK.Biodiv.Smpl` object, you will write code to make a series of histograms for *Distance to coast* [`Dis2Cost`] where you will change the number of brakes to:
  1) "Sturges" intervals.
  2) The 1st, 2nd, 3rd, and 4th quantiles.
  3) 15 intervals
  2) 30 intervals

</div>

```{r Normality2}
# Histograms for Distance to coast

## Sturges" intervals
hist(x = DK.Biodiv.Smpl$Dis2Cost, # Vector to summarize
     breaks = "Sturges" # how the brakes in the histogram are defined
     )

## quantile intervals
hist(x = DK.Biodiv.Smpl$Dis2Cost, # Vector to summarize
     breaks = quantile(na.omit(DK.Biodiv.Smpl$Dis2Cost)) # how the brakes in the histogram are defined
     )

## 15 intervals
hist(x = DK.Biodiv.Smpl$Dis2Cost, # Vector to summarize
     breaks = 15 # how the brakes in the histogram are defined
     )

## 30 intervals
hist(x = DK.Biodiv.Smpl$Dis2Cost, # Vector to summarize
     breaks = 30 # how the brakes in the histogram are defined
     )
```

<div class="alert alert-success">
**Question:** Based on these histograms, does *Distance to coast* appear to be normally distributed, left-skewed, or right-skewed?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
Based on all histograms is clear that *Distance to coast* is left-skewed (most values are small). 
</div>


<div class="alert alert-info">
**Your task:**

* Based on the data stored in the `DK.Biodiv.Smpl` object, you will write code to make a density plot for *Nature Density* [`NatDensBasMp`], *Distance to coast* [`Dis2Cost`], *Human Impact Index* [`HII`], *Slope* [`Slope30mAgg`],  and *Precipitation Seasonality* [`PrecSea`].

  * The function `density()` **MUST* be used to build the kernel density estimates.

  * You **MUST** use a `for()` loop** for making this series of plots.

  * All density plots need to be placed in the same plotting space divided into two (2) rows and three (3) columns.
 </div>
 
```{r Normality3}
# plot a density plots for env predictors in a 2x3 plotting space

## set the ploting space to a 2x3 space
par(mfrow = c(2,3))

# the loop
for (i in names(DK.Biodiv.Smpl)[4:8]){ # Loop over all the selected variables
  DensityTmp <- density(x = na.omit(DK.Biodiv.Smpl[,i]) #data from which the estimate is to be computed
                        )
  plot(DensityTmp, # Plot the kernel density estimates
       main = i # Add the name of the variable as a main title
  )
}  
```

<div class="alert alert-success">
**Question:** Based on these density plots, which variables **LOOK** normally distributed, left-skewed, or right-skewed?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
**Normally distributed:** HII, Slope30mAgg, and PrecSea
**Left-skewed:** Dis2Cost
**Right-skewed:** NatDensBasMp
</div>


<div class="alert alert-info">
**Your task:**

* Based on the data stored in the `DK.Biodiv.Smpl` object, you will test the normality of *Nature Density* [`NatDensBasMp`], *Distance to coast* [`Dis2Cost`], *Human Impact Index* [`HII`], *Slope* [`Slope30mAgg`],  and *Precipitation Seasonality* [`PrecSea`].

  * You will use **Q_Q plots** test for normality using the `qqnorm()` and `qqline()` functions.

  * You **MUST** use a `for()` loop to make all the tests.

  * All **Q_Q plots** need to be placed in the same plotting space divided into two (2) rows and three (3)

 </div>

```{r Normality4}
## set the ploting space to a 2x3 space
par(mfrow = c(2,3))

# the loop
for (i in names(DK.Biodiv.Smpl)[4:8]){ # Loop over all the selected variables
  qqnorm(na.omit(DK.Biodiv.Smpl[,i]), # Plot the QQ-plot
         main = paste("Normal Q-Q plot - ", i) # add a title with the variable name
         )
  qqline(na.omit(DK.Biodiv.Smpl[,i])) # add the trend line
}
```

<div class="alert alert-success">
**Question:** Based on these QQ-plots, which variables **LOOK** normally distributed?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
Both `HII` and `Slope30mAgg` LOOK normally distributed, as the theoretical quantiles and observed quantiles almost fall along the 1:1 line (solid diagonal line in the plot).
</div>

<div class="alert alert-info">
**Your task:**

* Based on the data stored in the `DK.Biodiv.Smpl` object, you will test the normality of *Nature Density* [`NatDensBasMp`], *Distance to coast* [`Dis2Cost`], *Human Impact Index* [`HII`], *Slope* [`Slope30mAgg`],  and *Precipitation Seasonality* [`PrecSea`].

  * You will use the `shapiro.test()` function to test for normality

  * You **MUST** use a `sapply()` function to make all the tests.

  * For each test you will summarize the Variable evaluated (`VarUsed`), the statistic value (`W`) with three (3) decimal points, and the p-value (`p.value`) with three (3) decimal points into one table.
 </div>
 
```{r Normality5}
# test the normality for evaluated variables using a shapiro.test()
ShapTest <- sapply(X = names(DK.Biodiv.Smpl)[4:8], # Vector defining the loop
                    FUN = function(x){ # Function to be applied over
                      # Shapiro test
                      ShpTestTmp <- shapiro.test(na.omit(DK.Biodiv.Smpl[,x]) # vector of data values
                                                 )
                      # Summary of Shapiro test
                      ShpTestOut <- data.frame(VarUsed = x, # Variable evaluated
                                               W = round(ShpTestTmp$statistic,3), # statistic value
                                               p.value = round(ShpTestTmp$p.value,3) # p-value
                                               )
                      return(ShpTestOut)
                    }
                    )
ShapTest
```
<div class="alert alert-success">
**Question:** Based on these density plots, which variables **ARE**  normally distributed?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
NONE of the evaluated variables is normaly distributed according to the Shapiro-test, as all these have p-values below 0.05; indicating a significant difference from a normal distribution. 
</div>

## *5. Transformations*

<div class="alert alert-info">
**Your task:**

* Based on the data stored in the `DK.Biodiv.Smpl` object, you will try different transformations for  *Nature Density* [`NatDensBasMp`] and test if these linearize the Q-Q plots.

  1) A `log10()` transformation.
  2) A squared-root (`sqrt()`) transformation.
  3) An Arc-sin (`asin()`) transformation.
  4) a Logit (`log(p/1-p)`) transformation.
</div>

```{r Trans1}
# Try different transformations for  *Nature Density
## set the ploting space to a 2x3 space
par(mfrow = c(2,3))

## NO transformation.
## plot the qq-pot and qqline
qqnorm(na.omit(DK.Biodiv.Smpl$NatDensBasMp), # Plot the QQ-plot
         main = "Normal Q-Q plot - NatDensBasMp" # add a title with the variable name
         )
qqline(na.omit(DK.Biodiv.Smpl$NatDensBasMp)) # add the trend line

## A `log10()` transformation.
log10.NatDensBasMp <- log10(DK.Biodiv.Smpl$NatDensBasMp)
## plot the qq-pot and qqline
qqnorm(na.omit(log10.NatDensBasMp), # Plot the QQ-plot
         main = "Normal Q-Q plot - log10.NatDensBasMp" # add a title with the variable name
         )
qqline(na.omit(log10.NatDensBasMp)) # add the trend line

## A squared-root (`sqrt()`) transformation.
Sqrt.NatDensBasMp <- sqrt(DK.Biodiv.Smpl$NatDensBasMp)
## plot the qq-pot and qqline
qqnorm(na.omit(Sqrt.NatDensBasMp), # Plot the QQ-plot
         main = "Normal Q-Q plot - Sqrt.NatDensBasMp" # add a title with the variable name
         )
qqline(na.omit(Sqrt.NatDensBasMp)) # add the trend line

## An Arc-sin (`asin()`) transformation.
Asin.NatDensBasMp <- asin(DK.Biodiv.Smpl$NatDensBasMp)
## plot the qq-pot and qqline
qqnorm(na.omit(Asin.NatDensBasMp), # Plot the QQ-plot
         main = "Normal Q-Q plot - Asin.NatDensBasMp" # add a title with the variable name
         )
qqline(na.omit(Asin.NatDensBasMp)) # add the trend line

## A Logit (`log(p/1-p)`) transformation.
Logit.NatDensBasMp <- log(DK.Biodiv.Smpl$NatDensBasMp/(1-DK.Biodiv.Smpl$NatDensBasMp))
## plot the qq-pot and qqline
qqnorm(na.omit(Logit.NatDensBasMp), # Plot the QQ-plot
         main = "Normal Q-Q plot - Asin.NatDensBasMp" # add a title with the variable name
         )
qqline(na.omit(Logit.NatDensBasMp)) # add the trend line

DK.Biodiv.Smpl$Logit.NatDensBasMp <- Logit.NatDensBasMp

```

<div class="alert alert-success">
**Question:** Based on these Q-Q plots, which transformation makes Nature Density more normally distributed?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
A Logit transformation makes Nature Density more normally distributed, as it is closer to the 1:1 line
</div>

<div class="alert alert-info">
**Your task:**

* Based on the data stored in the `DK.Biodiv.Smpl` object, you will try different transformations for  *Distance to coast* [`Dis2Cost`] and test if these linearise the Q-Q plots.

  1) A `log10()` transformation.
  2) A squared-root (`sqrt()`) transformation.
  
</div>

```{r Trans2}
# Try different transformations for  Distance to coast
## set the ploting space to a 2x3 space
par(mfrow = c(2,2))

## NO transformation.
## plot the qq-pot and qqline
qqnorm(na.omit(DK.Biodiv.Smpl$Dis2Cost), # Plot the QQ-plot
         main = "Normal Q-Q plot - Dis2Cost" # add a title with the variable name
         )
qqline(na.omit(DK.Biodiv.Smpl$Dis2Cost)) # add the trend line

## A `log10()` transformation.
log10.Dis2Cost <- log10(DK.Biodiv.Smpl$Dis2Cost)
## plot the qq-pot and qqline
qqnorm(na.omit(log10.Dis2Cost), # Plot the QQ-plot
         main = "Normal Q-Q plot - log10.Dis2Cost" # add a title with the variable name
         )
qqline(na.omit(log10.Dis2Cost)) # add the trend line

DK.Biodiv.Smpl$log10.Dis2Cost <- log10.Dis2Cost
## A squared-root (`sqrt()`) transformation.
Sqrt.Dis2Cost <- sqrt(DK.Biodiv.Smpl$Dis2Cost)
## plot the qq-pot and qqline
qqnorm(na.omit(Sqrt.Dis2Cost), # Plot the QQ-plot
         main = "Normal Q-Q plot - Sqrt.Dis2Cost" # add a title with the variable name
         )
qqline(na.omit(Sqrt.Dis2Cost)) # add the trend line

```

<div class="alert alert-success">
**Question:** Based on these Q-Q plots, which transformation makes Distance to coast more normally distributed?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
A Log transformation makes Distance to coast more normally distributed, as it is closer to the 1:1 line
</div>

## *6. Modelling*

<div class="alert alert-info">
**Your task:**

* Based on the data stored in the `DK.Biodiv.Smpl` object, you will evaluate the correlations between *Nature Density* [`NatDensBasMp`], *Distance to coast* [`Dis2Cost`], *Human Impact Index* [`HII`], *Slope* [`Slope30mAgg`],  and *Precipitation Seasonality* [`PrecSea`]. **Remember to use the appropriate correlation coefficient**.

</div>
```{r Model1}
# Evaluate the correlations between predictors
cor(x= na.omit(DK.Biodiv.Smpl[,names(DK.Biodiv.Smpl)[4:8]]), # matrix with values
    method = "spearman" # correlation coefficient to use
    )

```

<div class="alert alert-info">
**Your task:**

* Based on the data stored in the `DK.Biodiv.Smpl` object, you will visualize a series of scatter-plots showing the asociations between *Nature Density* [`NatDensBasMp`], *Distance to coast* [`Dis2Cost`], *Human Impact Index* [`HII`], *Slope* [`Slope30mAgg`],  and *Precipitation Seasonality* [`PrecSea`]. . 
  * For this you **MUST** use the function `pairs()` and try to add a smoothed line in each panel

</div>

```{r Model2}
# visualize the correlations between
pairs(DK.Biodiv.Smpl[,names(DK.Biodiv.Smpl)[4:8]], # What to plot
      panel = panel.smooth, # add a smoothed line in each panel
      cex = 1.5, pch = 19, bg = "black", # Set the point to filled, black and large points
      cex.labels = 2, font.labels = 2 # Set the size and boldface for the labels
      )

```

<div class="alert alert-success">
**Question:** DO you see strong associations between predictors? is yes/no, What does that indicate?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
There are only weak to moderate associations between predictors (most between -0.7 and 0.7). 
</div>


<div class="alert alert-info">
**Your task:**

* Based on the data stored in the `DK.Biodiv.Smpl` object, you will visualize the scatter plot showing the association between *Richness of all sampled species* [`S.AllGrp`] as the response variable (y-axis) and *Nature Density* [`NatDensBasMp`], *Distance to coast* [`Dis2Cost`], *Human Impact Index* [`HII`], *Slope* [`Slope30mAgg`],  and *Precipitation Seasonality* [`PrecSea`] as predictors (x-axis). 

  * All **scatter-plots** need to be placed in the same plotting space divided into two (2) rows and three (3)
  * Need to add the correlation coefficient with three (3) decimal points as a text insert on the top-right of each panel.
  * You most use a `for()` loop to make this plot.

</div>

```{r Model}
# Visualize the scatter plot showing the association between *Richness (y-axis) and all predictors (x-axis)
## set the ploting space to a 2x3 space
par(mfrow = c(2,3))

## Loop
for (i in names(DK.Biodiv.Smpl)[4:8]){ # Loop over all the selected variables
## Plot the scatter plot
  plot(DK.Biodiv.Smpl[,c(i,"S.AllGrp")], # data to plot
       pch = 19, # make filled circles
       main = i # add the variable name as a Title
       )
  ## Define the trend line regression
  LineTrnd <- line(DK.Biodiv.Smpl[,c(i,"S.AllGrp")]) # data to define the liner relation
  ## plot the trend line
  abline(coef(LineTrnd), # Coefficnets of the trend line
         lwd = 2, col ="red" # Parameter for the line (thick line and red)
         )
  ## Estimate the correlation 
  CorTest <- cor(na.omit(DK.Biodiv.Smpl[,c(i,"S.AllGrp")]),method="spearman")
  ## Add the correlation as a text insert
  legend("topright", # Put the legend in the top right corner
         legend = round(CorTest[2,1],3), # add the correlation coefficient
        bty = "n",
        cex=1.5)
}
```


<div class="alert alert-success">
**Question:** Based on this figure, which is the best predictor of species richness in the dataset?
</div>
    
<div class="alert alert-success"; style = "background-color : White; color: black">
Figure points to *Slope* and *Precipitation seasonality as the best predictors for species richness as these have the strongest correlations. 
</div>


## *7. Final Thoughts*

Exploratory data analysis is when a person used different approaches to learn about general patterns or tendencies in the data and find specific occurrences that deviate from the general patterns. Much like a detective explores a crime scene, collects evidence and draws conclusions, a statistician explores data using graphical displays and suitable summaries to draw conclusions about the main message of the data.

Here you have taken the first steps in this direction by describing the variation in the data, assessing the differences between groups, and establishing the relations between variables.