---
title: "DataExploration_Practical"
author: '"YOUR NAME HERE"'
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **The Set-up**

## *Learning objectives.*
1.	Load tables into `R` and explore their content.
2.  Execute basic looping routines using the `*apply()` and `for()` functions.
3.  Graphically explore the variability in the variables in a data set based.  
4.	Graphically compare the variables values between categorical groups  
5.	Establish the normality of a variable and evaluate different transformations to achieve normality.
6.	Determine the correlation between two random variables and the significance of such correlation.  


## *Before you start*
The practical will be run using *RStudio cloud* to avoid individual machine troubleshooting. This means you **need** to have an account for *RStudio cloud*.

However, remember to have [`R`](https://cran.r-project.org) and [R-studio](http://www.rstudio.com/) in your/room computer so you can work on the project after class.

**You should have also **
1. Read chapter 2 of Borcard et al. Numerical ecology with R. Springer, 2018.
2. Done the Tutorial for this week.

## *The way it's going to be run*

In this 3h practical you will to implement what you have learned in the class about Exploratory Data Analysis in `R`.

You will be building and **R-Markdown** document during the class. **R-Markdown** documents allows you to combine your code, its results, and your prose commentary into a single document. If you manage to produce a document your code works. The resulting file (a html) is what you will hand-in at the end of the practical.

This 3h practical is divided into two parts. **DO NOT COPY AND PASTE THE CODE!** 

Also try to write your own `R`-code and run it **DO NOT COPY THE CODE OF SOMEONE ELSE**. Writing your code and figuring out why it does not work is the best way to understand the dynamics of the `R` language and the tasks you are doing. Also, you have an excellent teaching assistant resource - ask him and/or me all the questions you have.

The final document will have for classes of boxes

<div class="alert alert-info">
<p class="text-center"; style = "font-size: 20px">
<strong>
The text in the blue filled box marked as **Your task:** states what you need to do</strong>
</strong>
</p>
</div>

```{r }
#
# The code in the window marked like this shows where you need to write your code
#

```

<div class="alert alert-success">
<p class="text-center"; style = "font-size: 20px">
<strong>The text in the green filled box marked as **question:** phrases a question.
</strong>
</p>
</div>


# **Exploring Denmark's biodiversity and environmental data.**

## *Load the Data*
<div class="alert alert-info">
**Your task:**

Load the data from **-------.csv*, located in the Web URL <https://raw.githubusercontent.com/AlejoOrdonez/StaGeoMod2022/-------/-------.csv>. For this:

* Define an object named URL.Loc with the string of the Web URL.

* Use `read.csv` to load the file -  The file is comma-separated 

* Save the object as `DK.Biodiv`.
</div>

```{r}
# Define an object named URL.Loc with the string of the Web URL.



# Read the file and save it as an object named `DK.Biodiv`




#

```

## *What is in the Table?*

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, write code to answer the flowing questions:

* What is the object class of `DK.Biodiv` data set?

* Which are the variable names in the `DK.Biodiv` data set?

* How many observations are in  the `DK.Biodiv` data set?

* List the variables that are `numeric`, `factors`, and `strings`.

</div>

```{r}
# What is the object class of `DK.Biodiv` data set?


#
```

```{r}
#Which are the variable names in the `DK.Biodiv` data set?


#
```

```{r}
#How many observations are in  the `DK.Biodiv` data set?


#
```

```{r}
#List the variables that are `numeric`, `factors`, and `strings`.


#
```

## *Summary statistics*

<div class="alert alert-info">
**Your task:**

Based on the data stored in the object `DK.Biodiv`, you will estimate.

* The mean of all `numerical` variables.

* The median of all `numerical` variables.

* The standard deviation of all `numerical` variables.



</div>

## *Graphical representations*

## *Normality tests*

## *Transformations*

## *Modelling*

## *Defining your hypotheses*
